# Materials and Methods {#sec-materials-and-methods}

## The Long-Term Phosphorus Fertilization Experiment {#sec-the-long-term-phosphorus-fertilization-experiment}

The soil samples for this thesis originate from a set of six long-term field trials in Switzerland, established by Agroscope between 1989 and 1992. The primary objective of these experiments was to validate and re-evaluate Swiss phosphorus (P) fertilization guidelines by assessing long-term crop yield responses to varying P inputs across different pedoclimatic conditions. A detailed description of the experimental design and site characteristics can be found in @hirteYieldResponseSoil2021.

The experiment was set up as a **completely randomized block design** with four field replications at each site. The core of the experiment consists of six fixed-plot treatments representing different P fertilization levels. These treatments were applied annually as triple superphosphate (TSP) before tillage and sowing. The application rates were strictly calibrated as percentages of the site- and crop-specific officially recommended P inputs (the Swiss GRUD norm): 0% (Zero), 33% (Deficit), 67% (Reduced), 100% (Norm), 133% (Elevated), and 167% (Surplus). This rigorous gradient ensures a massive divergence in cumulative P-balances across the 30-year timeframe, forcing distinct pedological responses.

Crucially, all other macronutrients—including Nitrogen (N), Potassium (K), and Magnesium (Mg)—were applied uniformly across all plots to prevent secondary yield limitations. However, it must be noted that these "non-limiting" applications were structurally calibrated to exactly 100% of their respective empirical Swiss GRUD recommendations, derived from the same matrix of operational soil tests and crop removal estimates criticized in this study. Consequently, the assumption that secondary nutrients were strictly non-limiting throughout the 30-year trial relies entirely on the epistemological validity of the empirical GRUD framework itself—a paradox that underscores the urgent need to transition toward mechanistic, site-specific nutrient modeling.

### Crop Yield, Phosphorus Uptake, and Balances

While the fundamental experimental design and baseline data collection follow the protocols described in @hirteYieldResponseSoil2021, the specific processing of the agronomic variables for this study diverges to explicitly isolate the chemical P-limitation. 

Crop yields were measured annually at harvest. Total fresh biomass was recorded, and representative subsamples were oven-dried to determine the absolute dry mass yield. Plant tissue phosphorus concentration ($C_P$) was subsequently quantified from these dried subsamples. The total phosphorus uptake ($P_{\text{uptake}}$) for each plot was calculated stoichiometrically as the product of the dry mass yield and the measured $C_P$.

To accurately quantify the historical legacy of the different fertilization treatments, the **Cumulative P Balance ($Cum\_P\_{\text{bal}}$)** was calculated for each individual plot. $Cum\_P\_{\text{bal}}$ is defined as the net sum of all P inputs (from the annual TSP fertilization) minus all P outputs (specifically, the P removed from the field via the harvested crop biomass) accumulated over the entire 30-year duration of the trial. This provides a direct, continuous metric of the anthropogenic P surplus or deficit, which drives the underlying thermodynamic buffering systems.

## Analytical Workflow and Data Integration

To answer the research questions, the methodology is structured into three highly synchronized, modular steps mapping the raw data to the final agronomic predictions:

```{tikz}
%| fig-cap: "Modular workflow synchronizing data sources, extraction methods, and mathematical models."
%| echo: false
%| fig-align: center
%| dev: pdf
%| fig-ext: pdf
%| engine.opts:
%|   extra.preamble: ["\\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, backgrounds, calc}"]

\begin{tikzpicture}[
    font=\sffamily\small,
    box/.style={rectangle, rounded corners, draw=black, thick, align=center, text width=3.5cm, minimum height=1cm},
    data/.style={cylinder, shape border rotate=90, aspect=0.25, draw=black, thick, fill=gray!10, text width=3cm, align=center, minimum height=1.2cm},
    arrow/.style={-{Stealth[scale=1.2]}, thick, draw=black!70},
    % Colors
    kinetics/.style={fill=cyan!10, draw=cyan!80!black},
    thermo/.style={fill=orange!10, draw=orange!80!black},
    bio/.style={fill=green!10, draw=green!80!black},
    eval/.style={fill=purple!10, draw=purple!80!black},
    % Group boxes
    groupbox/.style={rectangle, inner sep=20pt, inner ysep=25pt}
]

% --- Left Column: Data Sources ---
\node[data] (kin_data) at (0, 6.5) {Time-Series Data\\(Water Extraction)};
\node[data] (bio_data) at (0, 3.25) {Biological Data\\$Y_{\mathrm{rel}}$, $C_P$, $P_{\mathrm{up, rel}}$};
\node[data] (thermo_data) at (0, 0) {Equilibration Data\\($P_{\mathrm{AAE10}}$ vs $P_{\mathrm{CO}_2}$)};

% --- Middle Column: Sub-Models ---
\node[box, kinetics] (kin_model) at (4.5, 6.5) {Kinetic NLME Model\\(Desorption Rate)};
\node[box, thermo] (thermo_model) at (4.5, 0) {Thermodynamic LMER\\(Freundlich Isotherm)};

% --- Right Column: Coefficients & Biological Model ---
\node[box, kinetics] (kin_coef) at (9, 6.5) {Kinetic Coefficients\\$k, v_0$};
\node[box, eval] (mech_model) at (9, 3.25) {Mechanistic Models\\(Covariates: $v_0, 1/b$)};
\node[box, thermo] (thermo_coef) at (9, 0) {Buffer Coefficients\\$b, 1/b$};

% --- Far Right Column: Evaluation ---
\node[box, eval] (comparison) at (13.5, 3.25) {Model Evaluation\\\& Comparison};

% --- Edges ---
% Kinetics (Top Row)
\draw[arrow] (kin_data) -- (kin_model);
\draw[arrow] (kin_model) -- (kin_coef);
% Flow into biological (Down)
\draw[arrow] (kin_coef) -- (mech_model);

% Thermo (Bottom Row)
\draw[arrow] (thermo_data) -- (thermo_model);
\draw[arrow] (thermo_model) -- (thermo_coef);
% Flow into biological (Up)
\draw[arrow] (thermo_coef) -- (mech_model);

% Biological (Middle Row)
\draw[arrow] (bio_data) -- (mech_model);
\draw[arrow] (mech_model) -- (comparison);

% Backgrounds
\begin{scope}[on background layer]
    \node[groupbox, fit=(kin_data) (kin_model) (kin_coef)] (phase1) {};
    \node[anchor=north west, font=\sffamily\bfseries, color=gray!70!black, shift={(10pt,-8pt)}] at (phase1.north west) {Desorption Kinetics};

    \node[groupbox, fit=(bio_data) (mech_model) (comparison)] (phase3) {};
    \node[anchor=north west, font=\sffamily\bfseries, color=gray!70!black, shift={(10pt,-8pt)}] at (phase3.north west) {Biological Scaling \& Evaluation};

    \node[groupbox, fit=(thermo_data) (thermo_model) (thermo_coef)] (phase2) {};
    \node[anchor=north west, font=\sffamily\bfseries, color=gray!70!black, shift={(10pt,-8pt)}] at (phase2.north west) {Buffer Thermodynamics};
\end{scope}

\end{tikzpicture}

```

## Experimental Sites {#sec-experimental-sites}

The six experimental sites are located in the main crop-growing regions of Switzerland: **Rümlang-Altwi (ALT)**, **Cadenazzo (CAD)**, **Ellighausen (ELL)**, **Grabs (GRA)**, **Oensingen (OEN)**, and **Zurich-Reckenholz (REC)**. The key soil properties are summarized below.

```{r}
#| label: tbl-sites-corrected
#| tbl-cap: "Soil characteristics of the six long-term experimental sites. Data adapted from Hirte et al. (2021)."
#| echo: false
#| results: asis

sites_df_corrected <- data.frame(
  Site = c("ALT", "CAD", "ELL", "GRA", "OEN", "REC"),
  `Soil Type (WRB)` = c("Calcaric Cambisol", "Eutric Fluvisol", "Eutric Cambisol", "Calcaric Fluvisol", "Gleyic-calc. Cambisol", "Eutric Gleysol"),
  `Clay (%)` = c(22, 8, 33, 17, 37, 39),
  `Silt (%)` = c(30, 52, 36, 49, 31, 36),
  `Sand (%)` = c(48, 40, 31, 34, 32, 25),
  `Organic C (g/kg)` = c(21, 14, 23, 16, 24, 27),
  `pH (H2O)` = c(7.9, 6.3, 6.6, 8.3, 7.1, 7.4),
  check.names = FALSE
)

knitr::kable(sites_df_corrected)
```

Soil samples for this thesis were collected in the year 2022 from the topsoil layer (0-20 cm).

## Phosphorus Desorption Kinetics and Thermodynamics {#sec-phosphorus-desorption-kinetics}

The laboratory extraction of phosphorus was based on the sequential kinetic principles established by @flossmannExtractionMethodCharacterizing1982, subsequently modified to capture high-resolution temporal kinetics and corrected to reflect thermodynamic activity.

### Adapted Kinetic Protocol for This Study {#sec-adapted-kinetic-protocol-for-this-study}

To capture the desorption process with high temporal resolution while preserving the native soil matrix, a modified water-extraction protocol was employed:

1.  **Pre-washing to Remove Initial Soluble P**: 10 g of air-dried soil (sieved to < 2 mm) was suspended in 200 ml of deionized water and horizontally shaken for 60 minutes at 120 rpm. The suspension was centrifuged for 15 minutes at 4000 rpm, and the supernatant containing the readily soluble P was completely decanted. The P concentration of this discarded fraction was not quantified in this study, as the primary objective was to reset the soil solution to $P \approx 0$ to isolate the subsequent solid-phase desorption kinetics. No soil was lost during decantation due to the highly compact nature of the pellet.
2.  **Kinetic Extraction**: Because the remaining saturated soil pellet was highly compacted from centrifugation, achieving full resuspension required rigorous shaking, meaning the reactive surface area was not perfectly constant at the exact onset of the experiment. The pellet was rapidly resuspended in 200 ml of fresh deionized water to initiate the desorption process. The suspension was shaken continuously. To capture the nonlinear initial release phase, 2 ml subsamples were drawn at precisely eight time points: 2, 4, 10, 15, 20, 30, 45, and 60 minutes.
3.  **Analysis**: Each subsample was immediately filtered through a 0.2 $\mu$m membrane filter to halt desorption, and the stoichiometric concentration of orthophosphate ($c(t)$) was determined colorimetrically using the malachite green method [@VanVeldhoven1987Malachite].

### Temporal Stability of Amorphous Metal Oxides

Although the kinetic parameters ($k$) and geochemical structural variables ($Fe_{\text{ox}}$, $Al_{\text{ox}}$) were measured comprehensively in the year 2023, they were applied to the entire 26-year historical dataset. This is justified because oxalate-extractable iron and aluminum represent native, poorly-crystalline structural minerals rather than highly dynamic labile nutrient pools. Under constant arable management, these amorphous phases are widely recognized as pedologically stable over decadal timescales [@schwertmann1964differenzierung; @lookman1995geostatistical]. Thus, the 2023 measurements serve as robust, invariant site-means to characterize the fundamental sorption capacity of each experimental field.

## Statistical Analysis and Modeling {#sec-statistical-analysis}

All data processing, mathematical modeling, and visualization were conducted using the R programming language (v. 4.3.1) [@R-base]. Primary packages included `nlme` [@R-nlme] for non-linear kinetic modeling, `lme4` [@R-lme4] for linear mixed-effects modeling, and `deSolve` and `ReacTran` for partial differential equation solving.

### 1. Modeling of Desorption Kinetics and the Desorption Velocity ($v_0$)

To establish the foundational kinetic parameters, physical desorption curves were generated using a modified sequential extraction protocol based on @flossmannExtractionMethodCharacterizing1982. Soil samples were first pre-washed to discard readily soluble P, then resuspended in deionized water and continuously shaken. Subsamples were extracted across a high-resolution time series (2, 4, 10, 15, 20, 30, 45, and 60 minutes), and the orthophosphate concentration was determined colorimetrically using the malachite green method.

**Deriving the Desorption Rate Constant ($k$):**
A non-linear mixed-effects model (`nlme`) was fitted to these high-resolution extraction curves. The desorption process was modeled as a hyperbolic, asymptotic saturation curve derived from the exact solution of the first-order rate equation:

$$
c(t) = c_{\text{max}} \times (1 - e^{-k \times t'})
$$

Where $c(t)$ is the desorbed P concentration at time $t$, $c_{\text{max}}$ is the maximum size of the desorbable P pool (the asymptote), $k$ is the first-order desorption rate constant, and $t'$ is an adjusted time ($t_{\text{min}} + 3 \text{ min}$) representing the empirically determined handling and resuspension time delay required to break apart the centrifuged pellet before true, uniform first-order desorption commenced. Sample-specific deviations were managed with the nested random effect structure `c_max + k ~ 1 | site / plot`, allowing both parameters to vary freely across sites and individual field plots.

**Upscaling via Pedotransfer Function (PTF):**
Because this labor-intensive 60-minute kinetic extraction cannot be retroactively applied to the entire 30-year historical STYCS archive, a fundamental **Pedotransfer Function (PTF)** was constructed to predict the rate constant for missing soils and treatments. To ensure the upscaling process remained rigorously grounded in established mechanistic physics rather than arbitrary empirical correlations, the PTF was structured upon the Arrhenius equation—where reaction rates are driven by the competing activation energies of dominant sorption surfaces. Accordingly, a linear model (`lm`) was trained to predict $\ln(k)$ using explicitly pedogenetic predictor variables: the natural log of the ratio of amorphous aluminum to iron oxides ($\ln(Al_{\text{ox}} / Fe_{\text{ox}})$) and soil pH.

By anchoring the PTF to these inherently pedogenetic (treatment-independent) geochemical traits, we successfully extrapolated a predicted rate constant ($k_{pred}$) for every spatial and temporal plot in the dataset. 

From these predicted parameters, a novel proxy for agronomic availability, the **Dynamic Desorption Velocity ($v_0$)**, was defined mathematically as the product of the predicted rate constant and the historically measured static concentration pool:

$$
v_0 = k_{pred} \cdot P_{\text{CO}_2}
$$

### 2. Thermodynamic $Q(I)$ Isotherms and Buffering Capacity

To deduce the Phosphorus Buffering Capacity (PBC), a thermodynamic Quantity-Intensity ($Q/I$) framework was constructed. The $P_{\text{AAE10}}$ extraction was utilized as the standard mass Quantity ($Q$, $\text{mg P kg}^{-1}$), while the Davies-corrected $a_{\text{max}}$ was utilized as the native Intensity ($I$, $\mu\text{M}$). It must be acknowledged that relying on $P_{\text{AAE10}}$ as a proxy for absolute thermodynamic Quantity carries inherent mechanistic limitations. As demonstrated by isotopic exchange kinetics (IEK) studies, the extraction fluid's EDTA chelator actively dissolves non-exchangeable calcium phosphates—particularly in high-pH or high-CEC clays—leading to isotopic recovery rates as low as 25% [@demariaExchangeabilityPhosphateExtracted2005]. Furthermore, in heavily buffered clay soils, the chelator physically saturates when overwhelmed by exchangeable background cations. Once saturated, the test becomes entirely "blind" to any further thermodynamic $Q$, artificially collapsing the measured $\Delta Q$ parameter. 

The relationship between these variables across the varying fertilizer treatments at each site was parameterized using a log-linearized **Freundlich Isotherm** [@freundlich1906over], which is universally recognized for describing ion sorption on heterogeneous soil surfaces:

$$
Q = K_f \cdot I^{\frac{1}{n}}
$$

The apparent Phosphorus Buffering Capacity (PBC) for each specific site and state was then calculated analytically as the first derivative of the fitted Freundlich isotherm with respect to Intensity:

$$
\text{PBC} = \frac{dQ}{dI} = K_f \cdot \frac{1}{n} \cdot I^{\left(\frac{1}{n} - 1\right)}
$$

**Predicting Quantity via Pedotransfer Functions (PTF)**
To evaluate the practical scalability of these thermodynamic concepts, empirical Quantity ($P_{\text{AAE10}}$) was modeled directly from Intensity using a log-linear mixed-effects model (`lme`):

$$
\ln(Q) = \beta_0 + \beta_I \cdot \ln(I) + \sum (\beta_j \cdot Z_j) + \ln(I) \sum (\beta_{int,j} \cdot Z_j) + u_{site} + u_{plot|site}
$$

*   **The Covariates ($Z_j$):** Geochemical pedotransfer variables including soil fine texture (clay + silt), pH, $C_{\text{org}}$, and exchangeable background cations (Ca, Mg, K). These are mechanistically justified as they structurally determine the density and binding energy of the soil's sorption complexes (e.g., amorphous metal oxides and calcium carbonates).
*   **Interaction Terms:** The covariates directly interact with Intensity because the *slope* of the Q/I isotherm (the buffer power) is fundamentally altered by these intrinsic geochemical properties.
*   **Random Effects ($u_{year|site}$):** Nested temporal and spatial intercepts designed to rigorously absorb unmeasured macro-climatic shifts and temporal biological chaos (e.g. frost timing, pest pressure, radiation) across the different experimental years, preventing pseudo-replication.

### 3. Agronomic Normalization: Controlling Pedoclimatic Noise

To rigorously separate underlying P-limitation from the immense baseline yield variance caused by pedoclimatic shifts and crop genetics, all raw yield and P-uptake metrics were normalized to a biologically strict empirical ceiling. For every combination of site, year, and crop, the absolute maximum value observed across all fertilization treatments was assigned as $100\%$ capacity.

$$
Y_{\text{rel}} = \frac{Y_{plot}}{\max_{treatment}(Y)_{site, year, crop}}
$$

Unlike using a specific fertilization label (e.g., the P166 treatment average), this method guarantees that the empirical Mitscherlich asymptote of $1.0$ is strictly bounded. It ensures that the unconstrained biological ceiling for that specific environment is defined by the actual absolute maximum biomass it produced, structurally eliminating random negative constraints (e.g., localized compaction or pest damage) that might artificially depress a specific treatment average. This mathematical operation removes the variance in the absolute biological ceiling, strictly isolating the chemical P-limitation from the overarching temporal pedoclimatic noise.

### 4. Bayesian Modeling of Agronomic Outcomes

To test the core hypothesis and rigorously handle the structural heteroskedasticity inherent in agronomic field data, the biological response variables were modeled using Bayesian non-linear mixed-effects models via the `brms` package [@bürknerBrmsPackageBayesian2017]. Unlike frequentist approaches that often suffer from algorithmic convergence failures (singularities) when fitting complex asymptotes near biological boundaries, the Bayesian framework allows us to define mathematically rigorous, biologically grounded priors to constrain the parameter space. 

Due to the extreme non-linearity of the models—which multiply asymptotic kinetics by exponential environmental covariates across hundreds of random intercepts—the computational load was massive. The models were compiled and executed using the highly optimized `cmdstanr` backend [@gabry2023cmdstanr], enabling within-chain multi-threading to parallelize the Hamiltonian trajectories. All models utilized the No-U-Turn Sampler (NUTS) with four Markov Chain Monte Carlo (MCMC) chains, each running for 2000 iterations (1000 warmup), ensuring robust exploration of the posterior distributions. 

To aid clarity across the manuscript, the three core model architectures are explicitly labeled:

**1. The Yield Model ($B_{Yield}$)**
The agronomic yield response was modeled using the classical Mitscherlich equation. The explicit mathematical formulation replaces the static rate constant with a dynamic effective rate ($c_{\text{eff}}$):

$$
Y_{\text{rel}} = 1 - \exp\left(- c_{\text{eff}} \cdot (\text{STP} + E_{\text{base}}) \right)
$$

Where $c_{\text{eff}}$ is dynamically modulated by a set of fixed pedoclimatic covariates ($Z_i$) and nested spatial random effects ($u$):

$$
c_{\text{eff}} = (c_{\text{base\_crop}} + u_{site} + u_{plot|site}) \cdot \exp\left( \sum \beta_i \cdot Z_i \right)
$$

*   **The Diffusion Bottleneck ($\beta_{\text{invb}}$):** The inverse physical buffer power ($1/b$) acts as the primary transport penalty. Soils with high buffer power cannot replenish the root depletion zone fast enough via diffusion, functionally decreasing the agronomic efficiency of the measured STP.
*   **Speciation and Solubility ($\beta_{pH}$):** Soil pH fundamentally controls the speciation of orthophosphate in solution and drives precipitation dynamics, altering thermodynamic stability.
*   **Stoichiometric Co-Limitation ($\beta_{N}, \beta_{K}, \beta_{Mg}$):** Following Liebig's Law of the Minimum, the plant's ability to efficiently convert P into biomass is inherently capped if Nitrogen, Potassium, or Magnesium are limiting.
*   **Pedoclimatic Noise ($\beta_{Temp}, \beta_{Prec}$):** Temperature and precipitation anomalies dictate physical soil moisture (governing diffusion transport) and the basic biological metabolic rate.

**2. The Plant Uptake Model ($B_{Uptake}$)**
Actual mass-flux of P into the plant was modeled using a fractional Michaelis-Menten biological limit:

$$
Uptake_{\text{rel}} = \frac{U_{\text{eff}} \cdot \text{STP}}{K_{m} + \text{STP}}
$$

Where the effective maximum uptake capacity ($U_{\text{eff}}$) expands upon the base physiological limit ($U_{\text{base}}$) via critical covariates:

$$
U_{\text{eff}} = U_{\text{base}} + \beta_{\text{temp}} \cdot Z_{\text{temp}} + \beta_{N} \cdot Z_{N} + \beta_{v0} \cdot Z_{v0} + u_{site} + u_{year|site}
$$

*   **The Kinetics ($\beta_{v0}$):** The dynamic desorption velocity ($v_0$) directly increases maximal uptake capacity by continuously resupplying the root depletion zone.
*   **Crop-Specific Affinity ($K_m$):** The half-saturation constant ($K_m$) is modeled as a fixed effect grouped by *crop species*, reflecting the differing physiological root architectures and transporter affinities of maize, wheat, etc.
*   **Random Effects ($u_{year|site}$):** Nested `site/year` intercepts absorb the massive macro-climatic shifts (e.g., a regional drought year systemically lowering the overall $U_{\text{base}}$).

**3. The Tissue Concentration Model ($B_{C_P}$)**
To expose the physiological dynamics of tissue concentrations, the apparent $C_P$ was modeled empirically using a **Mitscherlich** saturation curve. Because true Piper-Steenbjerg carbon dilution penalties (a drop in concentration at initial growth stages) were statistically undetectable across the pedoclimatic noise, the phenomena clearly dictate a simpler biological reality: A crop-specific baseline tissue concentration ($C_{\text{min}}$), followed by an asymptotic **Luxury Accumulation** phase bounded by $\Delta C$:

$$
C_P = C_{\text{min}} + \Delta C \cdot \left(1 - \exp\left(- c_{\text{eff}} \cdot \text{STP}\right)\right)
$$

*   **Crop-Specific Baseline ($C_{\text{min}}$):** The fundamental, minimum tissue concentration required for the plant to physically exist, modeled as a fixed effect grouped by crop species (`~ crop - 1 + (1 | site/year)`).
*   **Luxury Accumulation ($\Delta C$):** Once basic metabolic needs are met, plants accumulate "luxury" P in their tissues up to a theoretical maximum.
*   **Accumulation Rate ($c_{\text{eff}}$):** This rate is significantly driven by pedoclimatic covariates and transport penalties, dictating how rapidly the soil can push excess phosphorus into the plant.

**Bayesian Prior Elicitation and Biological Bounds**
To prevent mathematical singularities and ensure the parameter space remained strictly within physical reality, informative bounds were placed on the fundamental biological constants:
- **Baseline and Maximum Yield ($Y_0, A$):** Bounded strictly $> 0$ with `normal(50, 50)` and `normal(100, 50)` respectively, preventing the MCMC algorithm from proposing mathematically impossible negative yields.
- **Maximum Uptake Velocity ($V_{\text{max}}$):** Bounded $> 0$ with `normal(30, 15)`, reflecting standard crop physiological limits (typically 15–45 kg P/ha).
- **Tissue Concentration Minimum ($C_{\text{min}}$):** Bounded $> 0$ with `normal(1.5, 1.0)`, as a baseline tissue concentration of $\approx 1.5 \text{ g/kg DM}$ is universally required for basic cellular function in these crops.
- **Environmental Covariates ($\beta$):** Strongly regularizing standard normal priors `normal(0, 1)` were applied to all standardized environmental $Z$-scores. This tight regularization was mathematically imperative; initial testing with weakly informative priors (`normal(0, 5)`) resulted in thousands of divergent transitions because the exponential parameterization ($\exp(\beta \cdot Z)$) caused rapid geometric explosions in the Hamiltonian gradients. Constraining the priors to $\pm 3$ standard deviations physically stabilized the MCMC sampler while allowing robust exploration of realistic agronomic effects.

```{r}
#| label: tbl-variables-updated
#| tbl-cap: "Comprehensive description of all response variables and covariates used in the thermodynamic and agronomic models, including their explicit preprocessing pipelines."
#| echo: false
#| results: asis

variables_df <- data.frame(
  Abbreviation = c(
    "$Y_{\\text{rel}}$", "$P_{\\text{up, rel}}$", "$\\ln(P_{\\text{AAE10}})$",
    "$P_{\\text{CO}_2}$", "$a_{\\text{CO}_2}$", "$k_{\\text{des}}$", "$v_{\\text{des}}$", "$\\text{PBC}^{-1}$",
    "$Fe_{\\text{ox}}, Al_{\\text{ox}}$", "$\\text{Clay} + \\text{Silt}$", "$\\text{pH}_{\\text{H}_2\\text{O}}$", "$C_{\\text{org}}$", "$Ca_{\\text{ex}}, Mg_{\\text{ex}}, K_{\\text{ex}}$",
    "$T_{\\text{anom}}, Pr_{\\text{anom}}$", "$N_{\\text{fert}}$"
  ),
  `Variable` = c(
    "Relative Yield", "Relative P Uptake", "Log Quantity Pool",
    "Empirical Intensity", "Thermodynamic Intensity", "Desorption Rate Constant", "Initial Desorption Velocity", "Inverse Buffer Power",
    "Amorphous Oxides", "Fine Texture", "Soil pH", "Organic Carbon", "Exchangeable Cations",
    "Climate Anomalies", "Nitrogen Fertilizer"
  ),
  Description = c(
    "Plot yield normalized by the local year's maximum.",
    "Plot P-uptake normalized by the local year's maximum.",
    "Standard mass of extractable P (AAE10 method).",
    "Water-extractable P ($CO_2$ saturated).",
    "Thermodynamic activity of the water pool.",
    "First-order rate constant of P desorption.",
    "Initial dynamic supply rate.",
    "Inverse of the dynamic $Q/I$ slope ($dI/dQ$).",
    "Amorphous Iron and Aluminum oxides.",
    "Sum of clay and silt fractions.",
    "Soil pH measured in water.",
    "Total organic carbon.",
    "AAE10-extractable background cations.",
    "Juvenile phase temp/prec deviations.",
    "Total applied nitrogen fertilizer."
  ),
  `Processing and Filtering` = c(
    "Normalized by the mean of the P166 treatment within the specific site $\\times$ crop $\\times$ year. Filtered for finite values.",
    "Normalized by the local P166 maximum. Rows with 0 uptake dropped.",
    "Natural log transformed to ensure strictly positive predictions and stabilize variance in mixed models.",
    "Raw measured values. Log-transformed for PTF training.",
    "Calculated via Davies equation accounting for ionic strength. Log-transformed.",
    "Missing historical plots were imputed using a pedogenetic PTF trained on log(Al/Fe) + pH. Standardized (z-score).",
    "Calculated as product of predicted $k$ and empirical $P_{\text{CO}_2}$. Standardized (z-score).",
    "Analytically derived from predicted Freundlich isotherms. Standardized (z-score).",
    "Averaged into invariant site-means to remove management artifacts, log-transformed, and standardized.",
    "Log-transformed sum to linearize bounds, standardized. Missing values imputed with site-means.",
    "Standardized directly.",
    "Log-transformed and standardized. Missing values imputed with site-means.",
    "Missing values imputed using site medians (or global medians if site missing). Log-transformed and standardized.",
    "Calculated as absolute deviation from the site's long-term historical mean. Standardized.",
    "Standardized directly."
  ),
  check.names = FALSE
)

# Apply styling to handle the long text in the processing column
knitr::kable(variables_df, format = "markdown", escape = FALSE)
```

### 5. Mechanistic Modeling of the Rhizosphere (Barber-Cushman Modification)

To mechanistically validate the field-scale statistical findings and visualize the limitations of static buffering capacity, a modified radial transport model based on Barber-Cushman principles [@barber1981soil] was constructed. Standard iterations of the Barber-Cushman model calculate radial diffusion using stoichiometric concentration ($C_l$) and a static buffer power ($b = \Delta Q / \Delta I$), effectively assuming instantaneous soil-solution equilibrium.

To reflect the kinetic bottlenecks identified in this study, the partial differential equation (PDE) for radial transport in cylindrical coordinates was modified. The driving gradient was kept as empirical concentration ($c_l$), and the static buffer power was replaced with a dynamic, first-order kinetic source term ($R_{\text{desorption}}$):

$$
\frac{\partial c_l}{\partial t} = D_e \left( \frac{\partial^2 c_l}{\partial r^2} + \frac{1}{r} \frac{\partial c_l}{\partial r} \right) + \frac{r_0 v_0}{r} \frac{\partial c_l}{\partial r} + R_{\text{desorption}}
$$

Where $D_e$ is the effective diffusion coefficient, $r$ is the radial distance, $r_0$ is the root radius, and $v_0$ is the water flux at the root surface. The kinetic source term resupplying the solution was defined using the experimentally derived parameters:

$$
R_{\text{desorption}} = k \cdot (c_{\text{max}} - c_l)
$$

Furthermore, the inner boundary condition at the root surface ($r = r_0$) was upgraded to reflect a dual-affinity Michaelis-Menten uptake system [@michaelis1913kinetik], accounting for both high-affinity (PHT1) and low-affinity (LATS) transporter kinetics. The binding kinetics of these transporters are governed by the fundamental thermodynamic driving force of the substrate: its chemical activity ($a$), not its total stoichiometric concentration ($c$) [@Atkins2011Physical; @Minton2001Macromolecular]. The boundary condition was defined as:

$$
J_r = \frac{V_{max1} \cdot (c_l - c_{\text{min}})}{K_{m1} + (c_l - c_{\text{min}})} + \frac{V_{max2} \cdot (c_l - c_{\text{min}})}{K_{m2} + (c_l - c_{\text{min}})}
$$

The modified PDEs were solved numerically in R utilizing the `ReacTran` and `deSolve` packages to generate dynamic spatial profiles of the P-depletion zone over time.

### 6. Derivation of Critical Thresholds ($P_{\text{crit}}$)

To translate the mechanistic models into actionable agronomic metrics, the critical Soil Test Phosphorus ($P_{\text{crit}}$) threshold was analytically derived from all three modeling frameworks. Rather than relying on arbitrary inflection points, each derivation represents a distinct physiological or agronomic paradigm:

**1. The Relative Yield Paradigm (The Production Target)**
Agronomically, $P_{\text{crit}}$ is classically defined as the STP required to achieve a specific operational threshold, commonly set as an $x\%$ relative yield target ($q_Y$, e.g., 0.95). By setting the relative yield target to $q_Y$ in the full Mitscherlich equation ($Y_{\text{rel}} = 1 - \exp(-c_{\text{eff}} \cdot (\text{STP} + E_{\text{base}}))$), the required soil concentration is extracted:
$$
P_{crit\_Y} = \frac{-\ln(1 - q_Y)}{c_{\text{eff}}} - E_{\text{base}}
$$
Where the effective rate constant incorporates the pedoclimatic penalties: $c_{\text{eff}} = c_{\text{base}} \cdot \exp(\beta_{kin} \cdot \text{Kinetics} + \dots)$.

**2. The Tissue Concentration Paradigm (The Physiological Buffer)**
Because true Piper-Steenbjerg minima were statistically absent, the tissue concentration critical limit was redefined as the threshold necessary to guarantee the plant's fundamental metabolic baseline ($C_{\text{base}}$) *plus* a luxury reserve ($\gamma$, representing an $x\%$ safety buffer). Solving the linear mixed-effects model ($C_P = C_{\text{base}} + S_{\text{acc}} \cdot \text{STP}$) for the target $C_{\text{base}}(1 + \gamma)$ yields:
$$
P_{crit\_C} = \frac{\gamma \cdot C_{\text{base}}}{S_{\text{base}} + \beta_{v0} \cdot v_0 + \beta_{\text{invb}} \cdot \frac{1}{b}}
$$
This derivation explicitly demonstrates that soils with a faster desorption velocity ($v_0$) or stronger buffer power ($1/b$) create a steeper accumulation slope ($S_{\text{acc}}$), inherently reducing the $P_{\text{crit}}$ required in the soil to safely buffer the crop against stress.

**3. The Relative Uptake Paradigm (The Mass-Transfer Limit)**
While yield integrates the entire season, instantaneous phosphorus uptake is strictly governed by Michaelis-Menten kinetics at the root surface. Defining a critical threshold as the STP required to reach an $x\%$ relative uptake target ($q_{\text{up}}$) yields a direct relationship with the effective half-saturation constant ($K_{\text{eff}}$):
$$
q_{\text{up}} = \frac{P_{crit\_up}}{K_{\text{eff}} + P_{crit\_up}} \implies P_{crit\_up} = \left(\frac{q_{\text{up}}}{1 - q_{\text{up}}}\right) K_{\text{eff}}
$$
Where $K_{\text{eff}}$ structurally expands upon the base crop affinity ($K_m$) via the same physical kinetic bottlenecks identified in the field data.