# Materials and Methods {#sec-materials-and-methods}

## The Long-Term Phosphorus Fertilization Experiment {#sec-the-long-term-phosphorus-fertilization-experiment}

The soil samples for this thesis originate from a set of six long-term field trials in Switzerland, established by Agroscope between 1989 and 1992. The primary objective of these experiments was to validate and re-evaluate Swiss phosphorus (P) fertilization guidelines by assessing long-term crop yield responses to varying P inputs across different pedoclimatic conditions. A detailed description of the experimental design and site characteristics can be found in @hirteYieldResponseSoil2021.

The experiment was set up as a **completely randomized block design** with four field replications at each site. The core of the experiment consists of six fixed-plot treatments representing different P fertilization levels, which were applied annually as superphosphate before tillage and sowing. These levels were based on percentages of the officially recommended P inputs: 0% (Zero), 33% (Deficit), 67% (Reduced), 100% (Norm), 133% (Elevated), and 167% (Surplus).

## Experimental Sites {#sec-experimental-sites}

The six experimental sites are located in the main crop-growing regions of Switzerland: **Rümlang-Altwi (ALT)**, **Cadenazzo (CAD)**, **Ellighausen (ELL)**, **Grabs (GRA)**, **Oensingen (OEN)**, and **Zurich-Reckenholz (REC)**. The key soil properties are summarized below.

```{r}
#| label: tbl-sites-corrected
#| tbl-cap: "Soil characteristics of the six long-term experimental sites. Data adapted from Hirte et al. (2021)."
#| echo: false

sites_df_corrected <- data.frame(
  Site = c("ALT", "CAD", "ELL", "GRA", "OEN", "REC"),
  `Soil Type (WRB)` = c("Calcaric Cambisol", "Eutric Fluvisol", "Eutric Cambisol", "Calcaric Fluvisol", "Gleyic-calc. Cambisol", "Eutric Gleysol"),
  `Clay (%)` = c(22, 8, 33, 17, 37, 39),
  `Sand (%)` = c(48, 40, 31, 34, 32, 25),
  `Organic C (g/kg)` = c(21, 14, 23, 16, 24, 27),
  `pH (H2O)` = c(7.9, 6.3, 6.6, 8.3, 7.1, 7.4),
  check.names = FALSE
)

knitr::kable(sites_df_corrected)
```

Soil samples for this thesis were collected in the year 2022 from the topsoil layer (0-20 cm).

## Phosphorus Desorption Kinetics and Thermodynamics {#sec-phosphorus-desorption-kinetics}

The laboratory extraction of phosphorus was based on the sequential kinetic principles established by @flossmannExtractionMethodCharacterizing1982, subsequently modified to capture high-resolution temporal kinetics and corrected to reflect true thermodynamic activity.

### Adapted Kinetic Protocol for This Study {#sec-adapted-kinetic-protocol-for-this-study}

To capture the desorption process with high temporal resolution while preserving the native soil matrix, a modified water-extraction protocol was employed:

1.  **Pre-washing to Remove Soluble P**: 10 g of air-dried soil was suspended in 200 ml of deionized water and shaken for 60 minutes at 120 Hz. The suspension was centrifuged for 15 minutes at 4000 rpm, and the supernatant containing the readily soluble P was discarded.
2.  **Kinetic Extraction**: The remaining soil pellet was resuspended in 200 ml of fresh deionized water. The suspension was shaken continuously, and subsamples were taken at eight time points: 2, 4, 10, 15, 20, 30, 45, and 60 minutes.
3.  **Analysis**: Subsamples were immediately filtered, and the stoichiometric concentration of orthophosphate ($c_i$) was determined colorimetrically using the malachite green method [@VanVeldhoven1987Malachite].

### Temporal Stability of Amorphous Metal Oxides

Although the kinetic parameters ($k$) and geochemical structural variables ($Fe_{ox}$, $Al_{ox}$) were measured comprehensively in the year 2023, they were applied to the entire 26-year historical dataset. This is justified because oxalate-extractable iron and aluminum represent native, poorly-crystalline structural minerals rather than highly dynamic labile nutrient pools. Under constant arable management, these amorphous phases are widely recognized as pedologically stable over decadal timescales [@schwertmann1964differenzierung; @lookman1995geostatistical]. Thus, the 2023 measurements serve as robust, invariant site-means to characterize the fundamental sorption capacity of each experimental field.

## Statistical Analysis and Modeling {#sec-statistical-analysis}

All data processing, mathematical modeling, and visualization were conducted using the R programming language (v. 4.3.1) [@R-base]. Primary packages included `nlme` [@R-nlme] for non-linear kinetic modeling, `lme4` [@R-lme4] for linear mixed-effects modeling, and `deSolve` and `ReacTran` for partial differential equation solving.

### 1. Modeling of Desorption Kinetics and the Desorption Velocity ($v_0$)

To derive the kinetic parameters, a non-linear mixed-effects model (`nlme`) was fitted to the exact solution of the first-order rate equation, utilizing the raw stoichiometric concentrations ($P_{CO_2}$):

$$c(t) = c_{max} \times (1 - e^{-k \times t'})$$

Where $c(t)$ is the P concentration at time $t$, $c_{max}$ is the maximum concentration of the desorbable P pool (the asymptote), $k$ is the first-order desorption rate constant, and $t'$ is an adjusted time ($t_{min} + 3 \text{ min}$) accounting for initial dissolution. Sample-specific deviations were modeled as random effects to capture the unique desorption characteristics of each soil.

Because the rate constant ($k$) is structurally invariant, it can be mathematically coupled with the historical Intensity data. From these parameters, a novel predictor for agronomic availability, the **Dynamic Desorption Velocity ($v_0$)**, was defined mathematically as the product of the rate constant and the historical static concentration:

$$v_0 = k \cdot P_{CO_2}$$

### 2. Thermodynamic $Q(I)$ Isotherms and Buffering Capacity

To deduce the Phosphorus Buffering Capacity (PBC), a thermodynamic Quantity-Intensity ($Q/I$) framework was constructed. The $P_{AAE10}$ extraction was utilized as the standard mass Quantity ($Q$, $\text{mg P kg}^{-1}$), while the Davies-corrected $a_{max}$ was utilized as the native Intensity ($I$, $\mu\text{M}$). 

The relationship between these variables across the varying fertilizer treatments at each site was parameterized using a log-linearized **Freundlich Isotherm**:

$$Q = K_f \cdot I^{\frac{1}{n}}$$

The true thermodynamic PBC for each specific site and state was then calculated analytically as the first derivative of the fitted Freundlich isotherm with respect to Intensity:

$$\text{PBC} = \frac{dQ}{dI} = K_f \cdot \frac{1}{n} \cdot I^{\left(\frac{1}{n} - 1\right)}$$

### 3. Agronomic Normalization: The Historical Baseline Strategy

To statistically evaluate the predictive power of the chemical parameters across six diverse pedoclimatic sites, raw crop yield and P-uptake data required robust normalization to isolate chemical P-limitation from overarching climatic noise. 

Standard normalization (dividing by the maximum yield of a given site within the specific experimental year) was rejected, as it artificially masks broad climatic limitations (e.g., drought) and forces the best plot of a poor year to equal 100%, removing the biological penalty. Instead, a **historical biological baseline strategy** was employed. 

Using the 40-year STYCS historical dataset, the true biological potential ($Y_{max}$ and $Uptake_{max}$) was calculated for each specific site and crop species by taking the 95th percentile of all fully fertilized (100% and 167% P) treatments. Current experimental yields and uptake values were divided by this historical baseline to calculate **Relative Historical Yield ($Y_{rel\_hist}$)**, providing a stable metric representing the percentage of the crop's long-term biological potential achieved.

### 4. Comparative Modeling of Agronomic Outcomes

To test the core hypothesis, two competing sets of linear mixed-effects models (`lmer`) were constructed to predict $Y_{rel\_hist}$ and Relative P-Uptake. Both sets shared an identical random effects structure—`(1|year) + (1|Site) + (1|Site:block)`—to absorb unmeasured pedoclimatic and meteorological variance.

* **Model 1: The Kinetic/Thermodynamic Approach:** Utilized the novel Activity Flux ($J_0$) as the primary fixed effect, testing the hypothesis that dynamic supply rate dictates plant availability.
* **Model 2: The Standard STP (GRUD) Approach:** Utilized the standard Swiss static soil tests ($P_{CO2}$ and $P_{AAE10}$) and their interaction as fixed effects.

```{r}
#| label: tbl-variables-updated
#| tbl-cap: "Description of key variables used in the thermodynamic and agronomic models."
#| echo: false

variables_df <- data.frame(
  Abbreviation = c("$Y_{rel\\_hist}$", "$Uptake_{rel}$", "$P_{bal}$", 
                   "$k$", "$c_{max}$", "$v_0$", 
                   "$P_{AAE10}$", "$\\text{PBC}$"),
  `Variable` = c("Relative Yield (Historic)", "Relative P Uptake", "Cumulative P Balance", 
                 "Rate Constant", "Maximum Concentration (Intensity)", "Desorption Velocity", 
                 "Standard Mass (Quantity)", "Phosphorus Buffering Capacity"),
  Unit = c("unitless", "unitless", "kg P ha⁻¹", 
           "min⁻¹", "mg P L⁻¹", "mg P L⁻¹ min⁻¹", 
           "mg P kg⁻¹", "mg kg⁻¹ (mg L⁻¹)⁻¹"),
  Description = c(
    "Plot yield normalized by the site/crop-specific 95th percentile of historical well-fertilized yields.",
    "Plot P-uptake normalized by historical maximum uptake for the specific site and crop.",
    "Long-term net P budget (inputs minus outputs).",
    "First-order rate constant of P desorption.",
    "Maximum empirical concentration of the water-desorbable P pool.",
    "Dynamic initial supply rate, calculated as $k \\times P_{CO2}$.",
    "Physical mass of extractable P (AAE10 method), acting as the Q-factor.",
    "The mathematical derivative ($dQ/dI$) of the site-specific Freundlich isotherm."
  )
)

knitr::kable(variables_df)
```

### 5. Mechanistic Modeling of the Rhizosphere (Barber-Cushman Modification)

To mechanistically validate the field-scale statistical findings and visualize the limitations of static buffering capacity, a modified radial transport model based on Barber-Cushman principles was constructed. Standard iterations of the Barber-Cushman model calculate radial diffusion using stoichiometric concentration ($C_l$) and a static buffer power ($b = \Delta Q / \Delta I$), effectively assuming instantaneous soil-solution equilibrium.

To reflect the kinetic bottlenecks identified in this study, the partial differential equation (PDE) for radial transport in cylindrical coordinates was modified. The driving gradient was kept as empirical concentration ($c_l$), and the static buffer power was replaced with a dynamic, first-order kinetic source term ($R_{desorption}$):

$$\frac{\partial c_l}{\partial t} = D_e \left( \frac{\partial^2 c_l}{\partial r^2} + \frac{1}{r} \frac{\partial c_l}{\partial r} \right) + \frac{r_0 v_0}{r} \frac{\partial c_l}{\partial r} + R_{desorption}$$

Where $D_e$ is the effective diffusion coefficient, $r$ is the radial distance, $r_0$ is the root radius, and $v_0$ is the water flux at the root surface. The kinetic source term resupplying the solution was defined using the experimentally derived parameters:

$$R_{desorption} = k \cdot (c_{max} - c_l)$$

Furthermore, the inner boundary condition at the root surface ($r = r_0$) was upgraded to reflect a dual-affinity Michaelis-Menten uptake system, accounting for both high-affinity (PHT1) and low-affinity (LATS) transporter kinetics:

$$J_r = \frac{V_{max1} \cdot (c_l - c_{min})}{K_{m1} + (c_l - c_{min})} + \frac{V_{max2} \cdot (c_l - c_{min})}{K_{m2} + (c_l - c_{min})}$$

The modified PDEs were solved numerically in R utilizing the `ReacTran` and `deSolve` packages to generate dynamic spatial profiles of the P-depletion zone over time.