\FloatBarrier

# Results
```{r setup-results}
#| include: false
# Meticulous Setup Chunk

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(brms)
  library(knitr)
  library(kableExtra)
  library(patchwork)
  library(latex2exp)
})

# Load cluster results payload
res <- readRDS("data/cluster_results.rds")
```

## The Blind Spot of Empirical Soil Tests

Standard empirical soil tests (STP) rely on chemical extraction to quantify a static "stock" of available phosphorus. However, these extractions are inherently blind to the dynamic physical constraints that govern how quickly that phosphorus can actually move through the soil matrix to the root surface. 

As demonstrated in @fig-kinetic-vs-empirical, soils that are classified identically by empirical extractions ($P_{CO_2}$, $P_{AAE10}$) can exhibit drastically different physical properties. A soil with a theoretically "adequate" $P_{CO_2}$ extraction value of 1.0 mg/kg can possess a diffusion restriction proxy ($1/b$) ranging anywhere from 0.05 to 0.45, and a desorption rate ($k$) that varies just as wildly. 

Ultimately, there appears to be no distinct connection between empirical STP measurements and fundamental kinetic variables. For any given STP value, a soil can exhibit a wide range of differing kinetic and diffusive properties, meaning these standard extractions do not inherently reflect the physical supply mechanisms of the soil matrix.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 9
#| fig-cap: "Distribution of long-term P balance across standard empirical tests ($P_{CO_2}$, $P_{AAE10}$) versus physical desorption kinetics (Buffer capacity $b$, Diffusion proxy $1/b$, and Desorption rate $k$). Red indicates a positive P balance (surplus), while green indicates a negative P balance (deficit). Note the lack of correlation between the empirical 'stock' and the physical 'flow' properties."
#| label: fig-kinetic-vs-empirical

D_Yield <- readRDS("models/D_Yield.rds")

d_plot <- D_Yield %>%
  select(site, plot_nr, year, inv_b_cons, k, soil_0_20_P_CO2, soil_0_20_P_AAE10, annual_P_balance) %>%
  distinct() %>%
  mutate(b = 1 / inv_b_cons) %>%
  filter(!is.na(b), !is.na(soil_0_20_P_CO2), !is.na(soil_0_20_P_AAE10), !is.na(annual_P_balance), inv_b_cons > 0, inv_b_cons < 0.5)

base_theme <- theme_bw() + theme(legend.position = "none")

# Row 1: b
p1 <- ggplot(d_plot, aes(x = soil_0_20_P_CO2, y = b, color = annual_P_balance, alpha = abs(annual_P_balance))) +
  geom_point(size = 3) +
  scale_color_gradient2(low = "green3", mid = "grey80", high = "red3", midpoint = 0) +
  scale_alpha_continuous(range = c(0.2, 1), guide = "none") +
  labs(x = tex_lab("$P_{CO_2}$ (mg/kg)"), y = tex_lab("Buffer Capacity ($b$)")) + base_theme

p2 <- ggplot(d_plot, aes(x = soil_0_20_P_AAE10, y = b, color = annual_P_balance, alpha = abs(annual_P_balance))) +
  geom_point(size = 3) +
  scale_color_gradient2(low = "green3", mid = "grey80", high = "red3", midpoint = 0) +
  scale_alpha_continuous(range = c(0.2, 1), guide = "none") +
  labs(x = tex_lab("$P_{AAE10}$ (mg/kg)"), y = tex_lab("Buffer Capacity ($b$)"), color = "P Balance") +
  theme_bw() # keep legend here for the patchwork

# Row 2: 1/b
p3 <- ggplot(d_plot, aes(x = soil_0_20_P_CO2, y = inv_b_cons, color = annual_P_balance, alpha = abs(annual_P_balance))) +
  geom_point(size = 3) +
  scale_color_gradient2(low = "green3", mid = "grey80", high = "red3", midpoint = 0) +
  scale_alpha_continuous(range = c(0.2, 1), guide = "none") +
  labs(x = tex_lab("$P_{CO_2}$ (mg/kg)"), y = tex_lab("Diffusion Proxy ($1/b$)")) + base_theme

p4 <- ggplot(d_plot, aes(x = soil_0_20_P_AAE10, y = inv_b_cons, color = annual_P_balance, alpha = abs(annual_P_balance))) +
  geom_point(size = 3) +
  scale_color_gradient2(low = "green3", mid = "grey80", high = "red3", midpoint = 0) +
  scale_alpha_continuous(range = c(0.2, 1), guide = "none") +
  labs(x = tex_lab("$P_{AAE10}$ (mg/kg)"), y = tex_lab("Diffusion Proxy ($1/b$)")) + base_theme

# Row 3: k
p5 <- ggplot(d_plot, aes(x = soil_0_20_P_CO2, y = k, color = annual_P_balance, alpha = abs(annual_P_balance))) +
  geom_point(size = 3) +
  scale_color_gradient2(low = "green3", mid = "grey80", high = "red3", midpoint = 0) +
  scale_alpha_continuous(range = c(0.2, 1), guide = "none") +
  labs(x = tex_lab("$P_{CO_2}$ (mg/kg)"), y = tex_lab("Desorption Rate ($k$)")) + base_theme

p6 <- ggplot(d_plot, aes(x = soil_0_20_P_AAE10, y = k, color = annual_P_balance, alpha = abs(annual_P_balance))) +
  geom_point(size = 3) +
  scale_color_gradient2(low = "green3", mid = "grey80", high = "red3", midpoint = 0) +
  scale_alpha_continuous(range = c(0.2, 1), guide = "none") +
  labs(x = tex_lab("$P_{AAE10}$ (mg/kg)"), y = tex_lab("Desorption Rate ($k$)")) + base_theme

# Combine via patchwork
(p1 + p2) / (p3 + p4) / (p5 + p6) + plot_layout(guides = "collect")
```

## Plant Response Variables

When evaluating predictive performance, it is important to strictly distinguish between crop yield and actual P uptake. While both serve as agronomic response variables, yield inherently exhibits greater variance and environmental noise. This instability occurs because the internal phosphorus concentration of the crop tissue ($C_P$) can fluctuate significantly—driven by luxury consumption in P-rich environments or physiological dilution during rapid biomass accumulation. Consequently, P uptake provides a more direct, mechanistically linked reflection of soil P availability than raw yield, which acts as a heavily buffered secondary response.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 4.5
#| fig-cap: "Total Yield versus Annual P Uptake grouped by crop type. The polynomial trend lines highlight the non-linear decoupling of yield from uptake at higher availability levels, driven by fluctuating internal plant P concentrations ($C_P$)."
#| label: fig-yield-vs-uptake

d_yield_uptake <- D_Yield %>%
  select(crop, annual_P_uptake, total_yield) %>%
  filter(!is.na(annual_P_uptake), !is.na(total_yield), !is.na(crop))

ggplot(d_yield_uptake, aes(x = annual_P_uptake, y = total_yield, color = crop)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, linewidth = 1.2) +
  labs(x = "P Uptake (kg P/ha)", y = "Total Yield (dt/ha)", color = "Crop") +
  theme_bw() +
  facet_wrap(~crop, scales = "free") +
  theme(legend.position = "none")
```

## Yield model comparison

To evaluate whether thermodynamic kinetic parameters offer an advantage over empirical supply classes, we compared the predictive power of five non-linear mixed-effects models against a null model. All models accounted for site-specific and crop-specific yield ceilings via hierarchical random effects ($1 | \text{site/year}$). They differed in how the base availability ($K_{base}$) of the STP ($P_{CO_2}$ or $P_{AAE10}$) was modulated. 


While the Heuristic $CO_2$ model achieved the lowest root mean square error (RMSE = 10.56) on the training data, the Bayesian leave-one-out cross-validation ($LOOIC$) indicated that this performance was associated with severe overfitting to the six pedoclimatic covariates ($p_{loo} = 1104.8$). 

In contrast, the Mechanistic model, which constrained $K_{base}$ strictly to the physical limits of the desorption rate ($k$) and buffer capacity ($1/b$), generalized better ($\Delta LOOIC = -1086$ vs. Heuristic $CO_2$; $p_{loo} = 247.2$). 

However, neither the empirical STPs nor the kinetic parameters demonstrated strong predictive power for absolute crop yield when compared to the Null model ($\Delta LOOIC = -1364$ vs. Mechanistic). This confirms that short-term agronomic yield is overwhelmingly dominated by local micro-climates and crop biology, rendering the specific physical P-supply mechanism secondary for short-term predictions.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: asis
library(knitr)

# Safely extract metrics
get_metric <- function(val, rnd = 2) {
  if (is.null(val)) return("-")
  sprintf(paste0("%.", rnd, "f"), val)
}

comp <- res$yield$comparison
looic_val <- function(m_name) {
  if (is.null(comp) || !(m_name %in% comp$model)) return("-")
  sprintf("%.0f", comp$looic[comp$model == m_name])
}
ploo_val <- function(m_name) {
  if (is.null(comp) || !(m_name %in% comp$model)) return("-")
  sprintf("%.1f", comp$p_loo[comp$model == m_name])
}
r2_cond <- function(r2_obj) {
  if (is.null(r2_obj)) return("-")
  sprintf("%.3f", r2_obj$conditional["R2", "Estimate"])
}
r2_marg <- function(r2_obj) {
  if (is.null(r2_obj)) return("-")
  sprintf("%.3f", r2_obj$marginal["R2", "Estimate"])
}

df_yield <- data.frame(
  Model = c("Base Model CO2", "Base Model AAE10", "Null Model", "Heuristic CO2", "Heuristic AAE10", "Mechanistic"),
  `Pedoclimatic Covariates` = c("(None)", "(None)", "(None)", "pH, Clay, Corg, Ca, Temp, Prec", "pH, Clay, Corg, Ca, Temp, Prec", "1/b, k, N, Temp, Prec"),
  RMSE = c(get_metric(res$yield$rmse$base), get_metric(res$yield$rmse$base_aae), get_metric(res$yield$rmse$null), get_metric(res$yield$rmse$heur), get_metric(res$yield$rmse$heur_aae), get_metric(res$yield$rmse$mech)),
  `R2_Marg` = c(r2_marg(res$yield$r2_base), r2_marg(res$yield$r2_base_aae), r2_marg(res$yield$r2_null), r2_marg(res$yield$r2_heur), r2_marg(res$yield$r2_heur_aae), r2_marg(res$yield$r2_mech)),
  `R2_Cond` = c(r2_cond(res$yield$r2_base), r2_cond(res$yield$r2_base_aae), r2_cond(res$yield$r2_null), r2_cond(res$yield$r2_heur), r2_cond(res$yield$r2_heur_aae), r2_cond(res$yield$r2_mech)),
  LOOIC = c(looic_val("mod_base_Y"), looic_val("mod_base_Y_aae"), looic_val("mod_null_Y"), looic_val("mod_heur_Y"), looic_val("mod_heur_Y_aae"), looic_val("mod_mech_Y")),
  p_loo = c(ploo_val("mod_base_Y"), ploo_val("mod_base_Y_aae"), ploo_val("mod_null_Y"), ploo_val("mod_heur_Y"), ploo_val("mod_heur_Y_aae"), ploo_val("mod_mech_Y")),
  check.names = FALSE
)

if (knitr::is_latex_output()) {
  library(kableExtra)
  out <- kable(df_yield, format = "latex", escape = FALSE, booktabs = TRUE, table.envir = "table*",
        caption = "Comparison of cross-validated predictive performance for yield models.\\label{tbl-yield-comparison}",
        col.names = c("Model", "Pedoclimatic Covariates", "RMSE", "$R^2_{Marg}$", "$R^2_{Cond}$", "LOOIC", "$p_{loo}$")) %>%
    kable_styling(position = "center") %>%
    row_spec(0, bold = TRUE)
  cat(out)
} else {
  cat("::: {#tbl-yield-comparison}\n\n")
  print(kable(df_yield, format = "markdown", col.names = c("Model", "Pedoclimatic Covariates", "RMSE", "$R^2_{Marg}$", "$R^2_{Cond}$", "LOOIC", "$p_{loo}$")))
  cat("\n\nTable: Comparison of cross-validated predictive performance for yield models.\n")
  cat(":::\n")
}
```

To evaluate true predictive power, we relied on the Expected Log Predictive Density (ELPD) calculated via Bayesian leave-one-out cross-validation (LOO-CV). Unlike in-sample error metrics such as RMSE, which invariably reward overly complex models, ELPD rigorously penalizes models that overfit to local noise. The difference in ELPD ($\Delta$ELPD) serves as a relative measure of out-of-sample performance. This metric fundamentally exposes models that achieve low training error merely by hallucinating relationships with non-causal covariates.

```{r}
#| label: fig-yield-looic
#| fig-cap: "Out-of-Sample Predictive Performance (LOO) of yield models relative to the best model. Error bars represent $\\pm 2$ standard errors of the ELPD difference."
#| warning: false
#| message: false
#| fig-width: 6
#| fig-height: 4.5

library(ggplot2)
library(dplyr)

res <- readRDS("data/cluster_results.rds")

# Extract LOO comparison
comp_df <- as.data.frame(res$yield$comparison)
comp_df$Model <- rownames(comp_df)
name_map <- c(
  "mod_mech_Y" = "Mechanistic",
  "mod_heur_Y" = "Heuristic CO2",
  "mod_heur_Y_aae" = "Heuristic AAE10",
  "mod_base_Y" = "Base CO2",
  "mod_base_Y_aae" = "Base AAE10",
  "mod_null_Y" = "Null Model"
)
comp_df$Model_Clean <- name_map[comp_df$Model]
comp_df$Model_Clean <- ifelse(is.na(comp_df$Model_Clean), comp_df$Model, comp_df$Model_Clean)

ggplot(comp_df, aes(x = reorder(Model_Clean, elpd_diff), y = elpd_diff)) +
  geom_point(size = 4, color = "#0072B2") +
  geom_errorbar(aes(ymin = elpd_diff - 2*se_diff, ymax = elpd_diff + 2*se_diff), width = 0.2, color = "#0072B2", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  theme_bw(base_size = 12) +
  labs(
    x = "Model",
    y = tex_lab("Difference in Expected Log Predictive Density ($\\Delta$ELPD)")
  )
```

```{r}
#| label: fig-yield-conditional
#| fig-cap: "Conditional effects of $P_{CO_2}$ STP on predicted yield across crops. The Mechanistic model (blue) produces constrained asymptotic responses bounded by physical desorption parameters. The Heuristic model (orange) demonstrates massive uncertainty and flat average responses, indicating that its predictive power was inappropriately shifted to pedoclimatic covariates."
#| warning: false
#| message: false
#| fig-width: 6
#| fig-height: 8

# Extract conditional effects
d_null <- res$yield$plot_data$null
d_heur <- res$yield$plot_data$heur
d_mech <- res$yield$plot_data$mech

if(!is.null(d_null)) d_null$Model <- "Null Model"
if(!is.null(d_heur)) d_heur$Model <- "Heuristic CO2"
if(!is.null(d_mech)) d_mech$Model <- "Mechanistic"

d_plot <- bind_rows(d_null, d_heur, d_mech)

ggplot(d_plot, aes(x = effect1__, y = estimate__, color = Model, fill = Model)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.2, color = NA) +
  facet_wrap(~crop, scales = "free", ncol = 2) +
  theme_bw(base_size = 12) +
  labs(
    x = tex_lab("STP $P_{CO_2}$ (mg / kg)"),
    y = "Predicted Yield (dt/ha)"
  ) +
  scale_color_manual(values = c("Null Model" = "gray50", "Heuristic CO2" = "#E69F00", "Mechanistic" = "#0072B2")) +
  scale_fill_manual(values = c("Null Model" = "gray50", "Heuristic CO2" = "#E69F00", "Mechanistic" = "#0072B2")) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-yield-shrinkage
#| fig-cap: "Posterior uncertainty (shrinkage) of the crop-specific base affinity ($K_{base}$) for the Mechanistic and Heuristic yield models. The massive 95% credible intervals for the Heuristic model indicate severe overfitting and an inability to constrain the fundamental agronomic relationship, whereas the Mechanistic model is tightly constrained by physical desorption kinetics."
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 5

extract_kbase <- function(df, model_name) {
  rn <- rownames(df)
  k_idx <- grep("^Kbase_crop", rn)
  sub_df <- df[k_idx, ]
  sub_df$Crop <- gsub("Kbase_crop", "", rn[k_idx])
  sub_df$Model <- model_name
  return(sub_df)
}

df_m <- extract_kbase(res$yield$parameters$mech, "Mechanistic")
df_h <- extract_kbase(res$yield$parameters$heur, "Heuristic CO2")

d_plot_k <- bind_rows(df_m, df_h)

ggplot(d_plot_k, aes(x = Crop, y = Estimate, color = Model)) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbar(aes(ymin = `l-95% CI`, ymax = `u-95% CI`), 
                position = position_dodge(width = 0.6), width = 0.2, linewidth = 1) +
  scale_y_log10(labels = scales::comma) +
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("Heuristic CO2" = "#E69F00", "Mechanistic" = "#0072B2")) +
  labs(
    x = "Crop",
    y = tex_lab("Posterior Estimate of Base Affinity ($K_{base}$) [Log Scale]")
  ) +
  theme(legend.position = "bottom")
```

## Phosphorus Uptake ($C_P$) model comparison

We repeated the identical model hierarchy to predict total plant phosphorus uptake rather than dry matter yield. 

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: asis

comp_u <- res$uptake$comparison
looic_val_u <- function(m_name) {
  if (is.null(comp_u) || !(m_name %in% comp_u$model)) return("-")
  sprintf("%.0f", comp_u$looic[comp_u$model == m_name])
}
ploo_val_u <- function(m_name) {
  if (is.null(comp_u) || !(m_name %in% comp_u$model)) return("-")
  sprintf("%.1f", comp_u$p_loo[comp_u$model == m_name])
}

df_uptake <- data.frame(
  Model = c("Base Model CO2", "Base Model AAE10", "Null Model", "Heuristic CO2", "Heuristic AAE10", "Mechanistic"),
  `Pedoclimatic Covariates` = c("(None)", "(None)", "(None)", "pH, Clay, Corg, Ca, Temp, Prec", "pH, Clay, Corg, Ca, Temp, Prec", "1/b, k, N, Temp, Prec"),
  RMSE = c(get_metric(res$uptake$rmse$base), get_metric(res$uptake$rmse$base_aae), get_metric(res$uptake$rmse$null), get_metric(res$uptake$rmse$heur), get_metric(res$uptake$rmse$heur_aae), get_metric(res$uptake$rmse$mech)),
  `R2_Marg` = c(r2_marg(res$uptake$r2_base), r2_marg(res$uptake$r2_base_aae), r2_marg(res$uptake$r2_null), r2_marg(res$uptake$r2_heur), r2_marg(res$uptake$r2_heur_aae), r2_marg(res$uptake$r2_mech)),
  `R2_Cond` = c(r2_cond(res$uptake$r2_base), r2_cond(res$uptake$r2_base_aae), r2_cond(res$uptake$r2_null), r2_cond(res$uptake$r2_heur), r2_cond(res$uptake$r2_heur_aae), r2_cond(res$uptake$r2_mech)),
  LOOIC = c(looic_val_u("mod_base_U"), looic_val_u("mod_base_U_aae"), looic_val_u("mod_null_U"), looic_val_u("mod_heur_U"), looic_val_u("mod_heur_U_aae"), looic_val_u("mod_mech_U")),
  p_loo = c(ploo_val_u("mod_base_U"), ploo_val_u("mod_base_U_aae"), ploo_val_u("mod_null_U"), ploo_val_u("mod_heur_U"), ploo_val_u("mod_heur_U_aae"), ploo_val_u("mod_mech_U")),
  check.names = FALSE
)

if (knitr::is_latex_output()) {
  library(kableExtra)
  out <- kable(df_uptake, format = "latex", escape = FALSE, booktabs = TRUE, table.envir = "table*",
        caption = "Comparison of cross-validated predictive performance for total phosphorus uptake models.\\label{tbl-uptake-comparison}",
        col.names = c("Model", "Pedoclimatic Covariates", "RMSE", "$R^2_{Marg}$", "$R^2_{Cond}$", "LOOIC", "$p_{loo}$")) %>%
    kable_styling(position = "center") %>%
    row_spec(0, bold = TRUE)
  cat(out)
} else {
  cat("::: {#tbl-uptake-comparison}\n\n")
  print(kable(df_uptake, format = "markdown", col.names = c("Model", "Pedoclimatic Covariates", "RMSE", "$R^2_{Marg}$", "$R^2_{Cond}$", "LOOIC", "$p_{loo}$")))
  cat("\n\nTable: Comparison of cross-validated predictive performance for total phosphorus uptake models.\n")
  cat(":::\n")
}
```

In a stark divergence from the yield models, the Heuristic $CO_2$ model demonstrated the highest predictive power both in-sample (RMSE = 0.62) and out-of-sample (LOOIC = 19374). 

Crucially, the Heuristic uptake model did not suffer from the severe overfitting observed in the yield predictions; its effective number of parameters ($p_{loo} = 163.2$) was nearly identical to the strictly constrained Mechanistic model ($p_{loo} = 160.7$). This indicates that the pedoclimatic covariates (pH, Clay, Temperature, Precipitation) were capturing true, generalizable physical signals rather than memorizing local noise.

This phenomenological divergence is biologically consistent. While absolute dry matter yield is a highly confounded metric capped by dozens of non-phosphorus environmental factors, luxury plant P-uptake ($C_P$) is tightly coupled to the physical diffusion of P to the roots and active root metabolism. Therefore, allowing local soil properties and climate variables to flexibly modulate the apparent availability of the empirical STP ($P_{CO_2}$) legitimately improves the prediction of biological P-acquisition over purely thermodynamic bounds.

```{r}
#| label: fig-uptake-looic
#| fig-cap: "Out-of-Sample Predictive Performance (LOO) of P-Uptake models. In stark contrast to dry matter yield, the Heuristic model genuinely outperforms the Mechanistic constraints out-of-sample."
#| warning: false
#| message: false

library(ggplot2)
res <- readRDS("data/cluster_results.rds")

comp_df <- as.data.frame(res$uptake$comparison)
comp_df$Model <- rownames(comp_df)
name_map <- c(
  "mod_mech_U" = "Mechanistic", "mod_heur_U" = "Heuristic CO2", 
  "mod_heur_U_aae" = "Heuristic AAE10", "mod_base_U" = "Base CO2", 
  "mod_base_U_aae" = "Base AAE10", "mod_null_U" = "Null Model"
)
comp_df$Model_Clean <- name_map[comp_df$Model]
comp_df$Model_Clean <- ifelse(is.na(comp_df$Model_Clean), comp_df$Model, comp_df$Model_Clean)

ggplot(comp_df, aes(x = reorder(Model_Clean, elpd_diff), y = elpd_diff)) +
  geom_point(size = 4, color = "#D55E00") +
  geom_errorbar(aes(ymin = elpd_diff - 2*se_diff, ymax = elpd_diff + 2*se_diff), width = 0.2, color = "#D55E00", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() + theme_bw(base_size = 12) +
  labs(x = "Model", y = tex_lab("Difference in Expected Log Predictive Density ($\\Delta$ELPD)"))
```

```{r}
#| label: fig-uptake-conditional
#| fig-cap: "Conditional effects of STP on P-Uptake. Because the pedoclimatic covariates capture true signal for P-acquisition, the Heuristic model produces stable, physically meaningful response curves without hallucinating."
#| warning: false
#| message: false
#| fig-width: 6
#| fig-height: 4.5

d_null <- res$uptake$plot_data$null
d_heur <- res$uptake$plot_data$heur
d_mech <- res$uptake$plot_data$mech

if(!is.null(d_null)) d_null$Model <- "Null Model"
if(!is.null(d_heur)) d_heur$Model <- "Heuristic CO2"
if(!is.null(d_mech)) d_mech$Model <- "Mechanistic"

d_plot <- dplyr::bind_rows(d_null, d_heur, d_mech)

ggplot(d_plot, aes(x = effect1__, y = estimate__, color = Model, fill = Model)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.2, color = NA) +
  facet_wrap(~crop, scales = "free") +
  theme_bw(base_size = 12) +
  labs(x = tex_lab("STP $P_{CO_2}$ (mg / kg)"), y = "Predicted P-Uptake (kg P/ha)") +
  scale_color_manual(values = c("Null Model" = "gray50", "Heuristic CO2" = "#E69F00", "Mechanistic" = "#0072B2")) +
  scale_fill_manual(values = c("Null Model" = "gray50", "Heuristic CO2" = "#E69F00", "Mechanistic" = "#0072B2")) +
  theme(legend.position = "bottom")
```

To fully understand *why* the pedoclimatic covariates induce severe overfitting in the yield models but provide genuine predictive power for P-uptake, we can examine the estimated effect sizes (and their uncertainties) across both agronomic responses (@fig-covariate-effects). 

```{r}
#| label: fig-covariate-effects
#| fig-cap: "Forest plot of estimated covariate modifiers on base P availability ($K_{base}$) for both Yield and P-Uptake. The Heuristic model's reliance on soil pH, clay, and organic carbon to predict yield exhibits massive uncertainty (wide 95% CIs) and instability, characteristic of overfitting. In contrast, the effects on P-Uptake are tightly constrained and biologically meaningful."
#| warning: false
#| message: false
#| fig-width: 6
#| fig-height: 4

library(ggplot2)
library(dplyr)
res <- readRDS("data/cluster_results.rds")

extract_covariates <- function(df, model_name, response_name) {
  rn <- rownames(df)
  cov_idx <- grep("^beta", rn)
  if(length(cov_idx) > 0) {
    sub_df <- df[cov_idx, ]
    sub_df$Parameter <- rn[cov_idx]
    sub_df$Parameter <- gsub("^beta|_Intercept", "", sub_df$Parameter)
    sub_df$Model <- model_name
    sub_df$Response <- response_name
    return(sub_df)
  }
  return(NULL)
}

df_yu_m <- extract_covariates(res$yield$parameters$mech, "Mechanistic", "Yield")
df_yu_h <- extract_covariates(res$yield$parameters$heur, "Heuristic CO2", "Yield")
df_up_m <- extract_covariates(res$uptake$parameters$mech, "Mechanistic", "Uptake")
df_up_h <- extract_covariates(res$uptake$parameters$heur, "Heuristic CO2", "Uptake")

d_plot <- bind_rows(df_yu_m, df_yu_h, df_up_m, df_up_h)

# Clean parameter names
d_plot$Parameter[d_plot$Parameter == "Temp"] <- "Temperature"
d_plot$Parameter[d_plot$Parameter == "Prec"] <- "Precipitation"
d_plot$Parameter[d_plot$Parameter == "Corg"] <- "C_{org}"
d_plot$Parameter[d_plot$Parameter == "N"] <- "Nitrogen"
d_plot$Parameter[d_plot$Parameter == "invb"] <- "1/b (Buffer)"
d_plot$Parameter[d_plot$Parameter == "k"] <- "k (Desorption)"

ggplot(d_plot, aes(x = Estimate, y = Parameter, color = Model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbar(aes(xmin = `l-95% CI`, xmax = `u-95% CI`), 
                position = position_dodge(width = 0.6), width = 0.2, linewidth = 1) +
  facet_wrap(~Response, scales = "free_x") +
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("Heuristic CO2" = "#E69F00", "Mechanistic" = "#0072B2")) +
  scale_y_discrete(labels = latex2exp::TeX) +
  labs(x = tex_lab("Effect Size on Base Availability (95% CI)"), y = "Covariate Modifier") +
  theme(legend.position = "bottom")
```
