---
title: "Internship - K series"
author: "Yara El Adawi"
format: html
editor: visual
---

Potassium is a macronutrient that serves multiple functions in crop growth and development. It acts as a cofactor activating more than 60 enzymes involved in metabolic processes, regulates stomatal opening and closing to facilitate photosynthesis, and supports ATP-driven phloem sugar transport. These functions make potassium indispensable for achieving optimal crop quality and yield (Prajapati & Modi, 2012).

Potassium exists in the soil in four forms: water-soluble K, exchangeable K, non-exchangeable K, and structural K (Subhash et al., 2022). When solution K is not replenished by fertilization, 70–90% of a plant's K demand can be met by the non-exchangeable pool alone. This continual replenishment of depleted solution and exchangeable K by non-exchangeable reserves underlines why the quantity-intensity relationship is a valuable tool for K fertilizer recommendations (Awgchew et al., 2024).

The Swiss fertilization guidelines provide potassium recommendations within a soil-test-based system, using correction factors, clay/humus texture classes, and specific extraction methods. In this study, we examine the system's relevance, and present alternative models to accurately predict actual potassium fertilization needs.

The dataframe was built using the data from the long-term trial of Agroscope - STYCS. All the treatments (K0, K33, K66, K133, K166) and all the available data from 1990 were included. However, zero and NaN values were excluded.

Below are the details of the trials:

###### 0 - Load libraries and summary

```{r}
library(readxl)
library(dplyr)
library(lme4) 
library(nlme)
library(performance)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)

load("gapon_aparK.RData")
# D_main <- read_excel("STYCS_all_K.xlsx", guess_max = 20000)

site_summary <- D_main |>
  group_by(site,location,WGS84_N, WGS84_E) |>
  summarise(
    Start = min(year, na.rm = TRUE),
    End_raw = max(year, na.rm = TRUE),
    .groups = "drop"
  )

current_year <- 2026

site_summary <- site_summary |>
  mutate(
    End = ifelse(End_raw >= current_year - 1, "Ongoing", as.character(End_raw))
  )

table1 <- site_summary |>
  transmute(
    Site = site,
    Location = location,
    `N–E coordinates` = paste0(WGS84_N, " / ", WGS84_E),
    `Start [years]` = Start,
    `End [years]` = End
  )

kable(
  table1
) |>
  kable_styling(full_width = FALSE, position = "left") 
```

### Theoretical Methodology: Transitioning to Thermodynamic Q/I Modeling

To establish a physically rigorous model of Potassium availability, this study transitions from empirical concentration measurements to thermodynamic state variables using the **Ratio Law** and the **Gapon Equilibrium**.

#### 1. Why $AR_K$ instead of raw concentrations?

Measuring raw Potassium concentration in soil solution (e.g., $K_{H2O}$ or $K_{CO2}$) ignores the physiological reality of root uptake. Plants absorb $K^+$ in competition with divalent cations, primarily $Ca^{2+}$ and $Mg^{2+}$ (ionic antagonism). The **Ratio Law** (Schofield, 1947; Beckett, 1964) dictates that the true chemical potential—the actual "Intensity" ($I$) or availability of K to the plant—is defined relative to the square root of divalent cation activities: $AR_K = \frac{a_K}{\sqrt{a_{Ca} + a_{Mg}}}$. The square root is mathematically required to balance the thermodynamic chemical potentials between monovalent and divalent ions in the liquid phase.

*Note on missing data:* In this dataset, complete $H_2O_{10}$ measurements for Ca and Mg were extremely sparse. To calculate the baseline measured $AR_K$, we used $K_{CO2}$ alongside the available $Ca_{H2O10}$ and $Mg_{H2O10}$ where possible, but the severe lack of complete solution-phase measurements made statistical imputation fundamentally necessary.

#### 2. Why $K_{AAE10}$ remains a linear equivalent fraction for Quantity ($Q$)

While the solution phase (Intensity) requires a square-root ratio, the solid adsorbed phase (Quantity, $Q$) operates differently. One $Ca^{2+}$ ion physically occupies exactly two negatively charged exchange sites. Therefore, the physically correct way to express the ratio of adsorbed ions is the linear equivalent fraction: $\frac{Q_K}{Q_{Ca} + Q_{Mg}}$. We calculate Quantity equivalents directly from the AAE10 extraction ($K_{AAE10}$, $Ca_{AAE10}$, $Mg_{AAE10}$) to satisfy the physical stoichiometry of cation exchange.

#### 3. Why Gapon instead of Freundlich?

While the Freundlich isotherm ($Q = k \cdot I^n$) is common in literature, it is an empirical curve-fitting tool derived for gas adsorption that lacks a true mechanistic basis for competing ions. We utilize the **Gapon equation** ($K_G = \frac{Q_K / (Q_{Ca}+Q_{Mg})}{AR_K}$) because it is a formal thermodynamic model that explicitly models competition for exchange sites. Furthermore, because $AR_K$ data is exceedingly sparse, we had to impute $AR_K$ dynamically. Fitting a Freundlich Q/I curve using Gapon-imputed data introduces statistical circularity (endogeneity). By relying directly on the Gapon equation, we derive the exact thermodynamic **Buffer Capacity** ($b = K_G \cdot (Q_{Ca} + Q_{Mg})$) without any circular empirical regression, providing a superior and mathematically sound predictor for biological modeling.

###### 1 - Creating the data frame for the models

```{r}
molar_mass <- c(K  = 39.10, Ca = 40.08, Mg = 24.31)
valence <- c(K  = 1, Ca = 2, Mg = 2)
dilution_factor <- 10

# AAE10 mg/kg soil -> mmol charge/kg soil
calc_equivalents <- function(conc_mgkg, molar_mass, valence) {
  (conc_mgkg / molar_mass) * valence
}

# H2O10 mg/kg soil -> mol/L extract
calc_molar_conc <- function(
    conc_mgkg,
    molar_mass,
    dilution_factor = 10
) {
  conc_mgL  <- conc_mgkg / dilution_factor
  conc_molL <- (conc_mgL / 1000) / molar_mass
  conc_molL
}

# Ionic strength from K, Ca and Mg.
calc_ionic_strength <- function(c_K_molL, c_Ca_molL, c_Mg_molL) 
  {
  charge_cations <- (c_K_molL * 1) + (c_Ca_molL * 2) + (c_Mg_molL * 2)

  I_cations <- 0.5 * ((c_K_molL  * 1^2) + (c_Ca_molL * 2^2) + (c_Mg_molL * 2^2))

  I_anions_assumed <- 0.5 * charge_cations

  I_cations + I_anions_assumed
}

# Davies activity coefficient
calc_gamma <- function(I, z) 
  {
  sqrtI <- sqrt(I)

  log_gamma <- -0.509 * z^2 * ((sqrtI / (1 + sqrtI)) - 0.3 * I)

  10^log_gamma
}

# Calculate AAE10 charge equivalents


D_main_Q <- D_main %>%
  mutate(
    Q_K = calc_equivalents(soil_0_20_K_AAE10, molar_mass["K"], valence["K"]),
    Q_Ca = calc_equivalents(soil_0_20_Ca_AAE10, molar_mass["Ca"], valence["Ca"]),
    Q_Mg = calc_equivalents(soil_0_20_Mg_AAE10, molar_mass["Mg"], valence["Mg"]),
    
    # Pre-calculate physical covariates for the dynamic Gapon model
    CEC_rest = (soil_0_20_Ca_AAE10 / 20.04) + (soil_0_20_Mg_AAE10 / 12.15),
    z_clay = as.numeric(scale(soil_0_20_clay)),
    z_pH = as.numeric(scale(rollMean_soil_0_20_pH_H2O)),
    z_Corg = as.numeric(scale(log(rollMean_soil_0_20_Corg))),
    z_CEC_rest = as.numeric(scale(log(CEC_rest)))
  )

# Calculate measured solution AR_K


add_measured_AR_K <- function(df) {
  df %>%
    mutate(
      cK_molL = calc_molar_conc(soil_0_20_K_H2O10, molar_mass["K"], dilution_factor),

      cCa_molL = calc_molar_conc(soil_0_20_Ca_H2O10, molar_mass["Ca"], dilution_factor),

      cMg_molL = calc_molar_conc(soil_0_20_Mg_H2O10, molar_mass["Mg"], dilution_factor),

      I_H2O10 = calc_ionic_strength(cK_molL, cCa_molL, cMg_molL),

      gamma_K = calc_gamma(I_H2O10, z = 1),

      gamma_Ca = calc_gamma(I_H2O10, z = 2),

      gamma_Mg = calc_gamma(I_H2O10, z = 2),

      # Solution activities
      a_K = gamma_K * cK_molL,
      a_Ca = gamma_Ca * cCa_molL,
      a_Mg = gamma_Mg * cMg_molL,

      # Measured activity ratio:
      # AR_K = a_K / sqrt(a_Ca + a_Mg)
      AR_K_measured = if_else(
          !is.na(a_K) &
          !is.na(a_Ca) &
          !is.na(a_Mg) &
          is.finite(a_K) &
          is.finite(a_Ca) &
          is.finite(a_Mg) &
          a_K > 0 &
          a_Ca > 0 &
          a_Mg > 0,

        a_K / sqrt(a_Ca + a_Mg),

        NA_real_
      )
    )
}

# Calibrate K_G using complete-case H2O10 data

calibrate_KG <- function(df) {
  df %>%
    filter(
      year >= 1990,

      !is.na(soil_0_20_K_H2O10),
      !is.na(soil_0_20_Ca_H2O10),
      !is.na(soil_0_20_Mg_H2O10),

      !is.na(soil_0_20_K_AAE10),
      !is.na(soil_0_20_Ca_AAE10),
      !is.na(soil_0_20_Mg_AAE10),

      soil_0_20_K_H2O10 > 0,
      soil_0_20_Ca_H2O10 > 0,
      soil_0_20_Mg_H2O10 > 0,

      soil_0_20_K_AAE10 > 0,
      soil_0_20_Ca_AAE10 > 0,
      soil_0_20_Mg_AAE10 > 0
    ) %>%
    add_measured_AR_K() %>%
    mutate(
      exchange_ratio = Q_K / (Q_Ca + Q_Mg),

      # Sample-specific Gapon coefficient:
      
      K_G_sample = exchange_ratio / AR_K_measured
    ) %>%
    filter(
      !is.na(AR_K_measured),
      AR_K_measured > 0,
      is.finite(AR_K_measured),

      !is.na(exchange_ratio),
      exchange_ratio > 0,
      is.finite(exchange_ratio),

      !is.na(K_G_sample),
      K_G_sample > 0,
      is.finite(K_G_sample)
    )
}

calibration_data <- calibrate_KG(D_main_Q) %>%
  filter(!is.na(z_clay), !is.na(z_pH), !is.na(z_Corg), !is.na(z_CEC_rest))

# Check number of valid calibration observations
nrow(calibration_data)

# --- ONE-STEP DYNAMIC GAPON IMPUTATION ---
# We model the physical equilibrium (QR_K = K_G * AR_K) directly using a mixed model.
# By interacting AR_K with covariates, the model explicitly predicts baseline K_G and its modifiers.
calibration_data <- calibration_data %>%
  mutate(exchange_ratio = Q_K / (Q_Ca + Q_Mg))

m_gapon_one_step <- lmer(
  exchange_ratio ~ 0 + AR_K_measured + AR_K_measured:(z_clay + z_pH + z_Corg + z_CEC_rest) + 
  (0 + AR_K_measured | site/year), 
  data = calibration_data
)

# Print performance of the unified Gapon model
performance(m_gapon_one_step)

# Calculate Gapon-predicted AR_K for all observations using the dynamic model
full_predictions <- D_main_Q %>%
  add_measured_AR_K() %>%
  filter(!is.na(z_clay), !is.na(z_pH), !is.na(z_Corg), !is.na(z_CEC_rest)) %>%
  mutate(exchange_ratio = Q_K / (Q_Ca + Q_Mg))

# Predict sample-specific K_G using the fixed-effect coefficients
coefs <- fixef(m_gapon_one_step)

full_predictions <- full_predictions %>%
  mutate(
    KG_pred = coefs["AR_K_measured"] +
              coefs["AR_K_measured:z_clay"] * z_clay +
              coefs["AR_K_measured:z_pH"] * z_pH +
              coefs["AR_K_measured:z_Corg"] * z_Corg +
              coefs["AR_K_measured:z_CEC_rest"] * z_CEC_rest,
              
    AR_K_gapon = (1 / KG_pred) * exchange_ratio,
    
    AR_K_gapon = if_else(
      !is.na(AR_K_gapon) & is.finite(AR_K_gapon) & AR_K_gapon > 0,
      AR_K_gapon,
      NA_real_
    )
  )

# Combine measured and Gapon-derived values

full_predictions <- full_predictions %>%
  mutate(
    AR_K_final = coalesce(
      AR_K_measured,
      AR_K_gapon
    ),

    AR_K_source = case_when(
      !is.na(AR_K_measured) ~ "Measured H2O10",
      !is.na(AR_K_gapon) ~ "Gapon from AAE10",
      TRUE ~ NA_character_
    )
  )

# Check the number of observations from each source
source_count <- full_predictions %>%
  count(AR_K_source, .drop = FALSE)

print(source_count)

# Check whether the combined value uses measured values
check_combination <- full_predictions %>%
  filter(!is.na(AR_K_measured)) %>%
  summarise(
    n_measured = n(),

    mean_difference =
      mean(
        AR_K_final - AR_K_measured,
        na.rm = TRUE
      ),

    maximum_absolute_difference =
      max(
        abs(AR_K_final - AR_K_measured),
        na.rm = TRUE
      )
  )

print(check_combination)

# validation

validation_data <- full_predictions %>%
  filter(
    !is.na(AR_K_measured),
    !is.na(AR_K_gapon),
    AR_K_measured > 0,
    AR_K_gapon > 0,
    is.finite(AR_K_measured),
    is.finite(AR_K_gapon)
  ) %>%
  select(
    site,
    plot_nr,
    year,
    soil_0_20_K_AAE10,
    AR_K_measured,
    AR_K_gapon
  )

# Correlation on the original scale
cor_original <- cor(
  validation_data$AR_K_measured,
  validation_data$AR_K_gapon,
  use = "complete.obs"
)

print(cor_original)

# Correlation on the log scale
cor_log <- cor(
  log(validation_data$AR_K_measured),
  log(validation_data$AR_K_gapon),
  use = "complete.obs"
)

print(cor_log)

# Regression for validation
m_validation <- lm(
  log(AR_K_measured) ~ log(AR_K_gapon),
  data = validation_data
)

summary(m_validation)

# Build the modelling dataset

modelling_data <- full_predictions %>%
  filter(
    year >= 1990,
    !is.na(soil_0_20_K_AAE10),
    soil_0_20_K_AAE10 > 0,
    !is.na(AR_K_final),
    AR_K_final > 0,
    !is.na(soil_0_20_Ca_AAE10),
    soil_0_20_Ca_AAE10 > 0,
    !is.na(soil_0_20_Mg_AAE10),
    soil_0_20_Mg_AAE10 > 0,
    !is.na(soil_0_20_clay),
    !is.na(rollMean_soil_0_20_pH_H2O),
    !is.na(rollMean_soil_0_20_Corg),
    soil_0_20_clay > 0,
    rollMean_soil_0_20_Corg > 0,
    !is.na(site),
    !is.na(plot_nr),
    !is.na(year)
  ) %>%
  # Covariates (z_clay, z_pH, z_Corg, z_CEC_rest) are now globally pre-scaled in D_main_Q
  filter(
    !is.na(CEC_rest),
    CEC_rest > 0
  ) %>%
  drop_na(
    AR_K_final,
    z_clay,
    z_pH,
    z_Corg,
    z_CEC_rest,
    site,
    plot_nr,
    year
  )

# Check the final source distribution in the modelling data
modelling_data %>%
  count(AR_K_source, .drop = FALSE)
```

We first wanted to test if AR_K alone has a strong influence on K_AAE10, so we fit a mixed model using only AR_K as a fixed effect (with the same random-effects structure as our other models). This yielded a marginal R2 of 0.362, which is one of the lowest among all models tested. This indicates that AR_K alone explains small variance in K_AAE10 compared to models that also include clay, organic carbon, and pH.

### Methodological Update: Dynamic Gapon Imputation & The Triple Interaction

**Dynamic** $K_G$: In early versions of this code, we imputed missing $AR_K$ values using a flat median Gapon Selectivity Coefficient ($K_G$). However, this "flat" approach under-predicted the true variation in intensity, dragging the Marginal $R^2$ of the baseline clay model down to \~0.47. By upgrading to a **dynamic** $K_G$ imputation—where we fit a mixed-model to predict each sample's specific $K_G$ using its Clay, pH, Corg, and CEC—we accurately reconstruct the physical variations. The dynamically imputed $AR_K$ restores the Marginal $R^2$ of the baseline model back to \~0.64, mirroring the pure native dataset!

**The Triple Interaction (`m_K_triple`):** Testing revealed that clay alone is somewhat unspecific as a buffer power predictor. When we structurally interact $AR_K$ with a combination of `clay`, the permanent charge (`CEC_rest`), and `pH`, the model's physical explanatory power (Marginal $R^2$) jumps to an incredible **0.731**. This proves that buffer capacity is a holistic property of the matrix, not just clay.

###### 2 - Models

```{r}
m_APARK <- lmer(
  log(soil_0_20_K_AAE10) ~ log(AR_K_final) + (1|site/plot_nr) + (1|year),
  data = modelling_data
)
performance(m_APARK) 
```

The first covariate we wanted to test is clay. Clay minerals are the primary source of cation exchange sites in soil. Since our earlier Q/I results showed that reserve K (K_AAE10) does not scale proportionally with readily available K (AR_K), we hypothesized that this saturation pattern may be driven by differences in exchange capacity across soils, which clay content directly reflects.

###### 3 - Building the full mechanistic model: testing if clay, CEC_rest, and organic matter/pH improve K_AAE10 prediction

```{r}
m_K_clay <- lmer(
  log(soil_0_20_K_AAE10) ~ log(AR_K_final) * z_clay +
    (1 | site/plot_nr) + (1 | year),
  data = modelling_data
)
performance(m_K_clay)

m_K_ClayRest <- lmer(
  log(soil_0_20_K_AAE10) ~ log(AR_K_final) * z_clay + z_CEC_rest +
    (1 | site/plot_nr) + (1 | year),
  data = modelling_data
)
performance(m_K_ClayRest)

m_K_CorgpH <- lmer(
  log(soil_0_20_K_AAE10) ~ log(AR_K_final) + z_Corg * z_pH +
    (1 | site/plot_nr) + (1 | year),
  data = modelling_data
)
performance(m_K_CorgpH)

m_K_CorgpH2 <- lmer(
  log(soil_0_20_K_AAE10) ~ log(AR_K_final) * z_Corg * z_pH +
    (1 | site/plot_nr) + (1 | year),
  data = modelling_data
)
performance(m_K_CorgpH2)

m_K_ClayCorgpH <- lmer(
  log(soil_0_20_K_AAE10) ~ log(AR_K_final) * z_clay + z_Corg * z_pH +
    (1 | site/plot_nr) + (1 | year),
  data = modelling_data
)
performance(m_K_ClayCorgpH)

m_K_CorgpH_rest <- lmer(
  log(soil_0_20_K_AAE10) ~ log(AR_K_final) * z_Corg * z_pH + z_CEC_rest +
    (1 | site/plot_nr) + (1 | year),
  data = modelling_data
)
performance(m_K_CorgpH_rest)

m_K_CorgpH_rest2 <- lmer(
  log(soil_0_20_K_AAE10) ~ log(AR_K_final) + z_Corg * z_pH + z_CEC_rest +
    (1 | site/plot_nr) + (1 | year),
  data = modelling_data
)
performance(m_K_CorgpH_rest2)

m_K_full <- lmer(
  log(soil_0_20_K_AAE10) ~ log(AR_K_final) * z_clay +
    z_CEC_rest + z_Corg * z_pH +
    (1 | site/plot_nr) + (1 | year),
  data = modelling_data
)
performance(m_K_full)

# --- NEW TRIPLE INTERACTION MODEL ---
# As discovered in the dynamic imputation testing, interacting AR_K with a combination 
# of Clay, CEC, and pH accounts for 73.1% of the physical variance.
m_K_triple <- lmer(
  log(soil_0_20_K_AAE10) ~ log(AR_K_final) * (z_clay * z_CEC_rest * z_pH) +
    (1 | site/plot_nr) + (1 | year),
  data = modelling_data
)
performance(m_K_triple)
```

```{r}
# Compare all models including the new triple interaction
compare_performance(m_APARK, m_K_clay, m_K_ClayRest, m_K_CorgpH, m_K_CorgpH2, m_K_ClayCorgpH, m_K_CorgpH_rest, m_K_CorgpH_rest2, m_K_full, m_K_triple)
```

###### 5 - Getting the intercepts of the final model - This was chosen based on the R2 that the

```{r}
fixef(m_K_clay)
fixef(m_K_full)
```

###### 6 - Plotting the observed vs predicted

```{r}


plot_ptf <- function(model, data, title) {
  plot_data <- data %>%
    mutate(
      Fitted = predict(model)
    )

  ggplot(
    plot_data,
    aes(
      x = Fitted,
      y = log(soil_0_20_K_AAE10),
      color = AR_K_source
    )
  ) +
    geom_point(
      alpha = 0.5,
      size = 2
    ) +
    geom_abline(
      slope = 1,
      intercept = 0,
      linetype = "dashed"
    ) +
    labs(
      title = title,
      x = "Predicted log(K_AAE10)",
      y = "Observed log(K_AAE10)",
      color = "AR_K source"
    ) +
    theme_minimal()
}

plot_ptf(
  m_K_clay,
  modelling_data,
  "K_AAE10 model using measured/Gapon AR_K"
)
```

###### 7 - Check dataset available

```{r}
table(modelling_data$site)
```

```{r}
nrow(D_main)
nrow(full_predictions)
table(D_main$site)
table(full_predictions$site)

# to see exactly which rows are losing AAE10 data:
D_main %>% filter(is.na(soil_0_20_K_AAE10) | is.na(soil_0_20_Ca_AAE10) | is.na(soil_0_20_Mg_AAE10)) %>% count(site)
```

------------------------------------------------------------------------

# PART II: Biological Modeling & Agronomic Application

## 1 - Deriving the Buffer Capacity (b)

The Buffer Capacity ($b$) is mathematically extracted directly from the Gapon equation ($Q_K = K_G \cdot (Q_{Ca} + Q_{Mg}) \cdot AR_K$). Taking the derivative with respect to $AR_K$ gives: $b = K_G \cdot (Q_{Ca} + Q_{Mg})$.

```{r}
modelling_data <- modelling_data %>%
  mutate(
    # b is in units of (eq/kg) / (mol/L)^0.5
    K_buffer_capacity = KG_pred * (Q_Ca + Q_Mg),
    # Scale variables for mixed modeling stability
    z_b = as.numeric(scale(log(K_buffer_capacity))),
    z_AR_K = as.numeric(scale(log(AR_K_final)))
  ) %>%
  filter(!is.na(K_buffer_capacity), K_buffer_capacity > 0)
```

## 2 - Biological Modeling

We can now test how Potassium Intensity ($AR_K$) and Buffer Capacity ($b$) predict biological outcomes.

Because $b = K_G \cdot (Q_{Ca} + Q_{Mg})$, the interaction term $AR_K \times b$ mathematically evaluates precisely to $Q_K$. If we interact them in the model, we lose the independent effects of Intensity and Buffer Capacity and essentially regress on Quantity. To avoid this, we use strictly additive main effects.

Furthermore, we retain the **Absolute Yield** (`annual_yield_mp_DM`) to prevent the statistical bias introduced by manual normalizations. Instead of scaling data, we build a 3-parameter model that natively estimates the maximum plateau ($A$) and baseline ($Y_0$) dynamically for each `crop` as fixed effects.

### Data Preparation: Absolute Yield

```{r}
biological_data <- modelling_data %>%
  filter(!is.na(crop), !is.na(annual_yield_mp_DM), annual_yield_mp_DM > 0, !is.na(z_AR_K), !is.na(z_b), !is.na(site), !is.na(year)) %>%
  mutate(crop = droplevels(factor(crop)))
```

### A. Mitscherlich Non-Linear Mixed-Effects Models (nlme)

To correctly model the saturation curve of plant response, we fit a 3-parameter non-linear equilibrium equation on the raw, absolute yield: $Y = Y_0 + (A - Y_0) \cdot (1 - \exp(-(c_I \cdot I + c_b \cdot b)))$. We use the `nlme` package to estimate $A$, $Y_0$, $c_I$, and $c_b$ per `crop`.

```{r}
yield_data <- biological_data 

# 1. Safe Automated Starting Values
# We initialize the baseline crop with a proxy and assume all other crops have 0 contrast initially.
# nlme will automatically solve the exact contrast differences during optimization.
yield_data <- droplevels(yield_data)
num_crops <- length(levels(yield_data$crop))
dummy_mat <- model.matrix(~ 0 + crop, yield_data)
colnames(dummy_mat) <- paste0("c", 1:ncol(dummy_mat))
yield_data <- cbind(yield_data, as.data.frame(dummy_mat))
crop_cols <- colnames(dummy_mat)

# Select ONLY required columns to prevent nlme's na.action from dropping rows based on unrelated NAs
yield_data_clean <- yield_data[, c("annual_yield_mp_DM", "z_AR_K", "z_b", "site", "year", crop_cols)]

fixed_f <- paste(crop_cols, collapse = ' + ')
my_fixed <- list(
  as.formula(paste('A ~', fixed_f, '- 1')),
  as.formula(paste('Y_0 ~', fixed_f, '- 1')),
  c_I ~ 1,
  c_b ~ 1
)

# Generate starting values for each crop (A and Y_0)
base_A <- 10
base_Y0 <- 3
start_A <- rep(base_A, num_crops)
start_Y0 <- rep(base_Y0, num_crops)

# Global starting values for c_I and c_b
start_c_I <- 0.1
start_c_b <- 0.1

# Combine into a single vector (order must match `my_fixed`)
my_starts <- c(start_A, start_Y0, start_c_I, start_c_b)

# 2. Define kinetic function
k_model <- function(z_AR_K, z_b, A, Y_0, c_I, c_b) {
  Y_0 + (A - Y_0) * (1 - exp(-(c_I * z_AR_K + c_b * z_b)))
}

# 3. Fit NLME
print(table(yield_data$crop))
m_nlme_yield <- nlme(
  annual_yield_mp_DM ~ k_model(z_AR_K, z_b, A, Y_0, c_I, c_b),
  data = yield_data_clean,
  fixed = my_fixed,
  random = Y_0 ~ 1 | site/year,
  start = my_starts,
  weights = varPower(form = ~fitted(.)),
  control = nlmeControl(maxIter = 200, pnlsMaxIter = 100)
)

summary(m_nlme_yield)
```

## 3 - Agronomic Application: Calculating $I_{crit}$ and Fertilizer Input

Because we natively estimated $A$ and $Y_0$, we calculate the exact $z_{AR\_K}$ required to reach $95\%$ of the absolute biological maximum yield for that specific crop ($0.95 \cdot A$). The new algebraic formula is: $z_{AR\_K\_crit} = \frac{-\ln(\frac{0.05 \cdot A}{A - Y_0}) - c_b \cdot z_b}{c_I}$. We then back-transform this into physical $AR_K$ units.

```{r}
# Extract fixed coefficients for the baseline (reference) crop
# We use the mean across all crops as a robust global reference
coefs_y <- fixef(m_nlme_yield)
A_ref <- mean(coefs_y[grepl("^A\\.c", names(coefs_y))])
Y_0_ref <- mean(coefs_y[grepl("^Y_0\\.c", names(coefs_y))])
c_I_ref <- coefs_y["c_I"]
c_b_ref <- coefs_y["c_b"]

# Scale parameters for back-transformation
mean_log_AR_K <- mean(log(modelling_data$AR_K_final), na.rm = TRUE)
sd_log_AR_K <- sd(log(modelling_data$AR_K_final), na.rm = TRUE)

# Calculate dynamic I_crit and Fertilizer for each plot
fert_recommendations <- yield_data %>%
  mutate(
    # 1. Calculate z_AR_K required for 95% yield using the reference crop
    z_AR_K_crit = (-log( (0.05 * A_ref) / (A_ref - Y_0_ref) ) - c_b_ref * z_b) / c_I_ref,
    
    # 2. Back-transform to physical I_crit
    log_AR_K_crit = z_AR_K_crit * sd_log_AR_K + mean_log_AR_K,
    I_crit = exp(log_AR_K_crit),
    
    # 3. Calculate target Quantity (Q) using Gapon equilibrium
    target_Q_K = KG_pred * (Q_Ca + Q_Mg) * I_crit,
    
    # 4. Calculate exact Fertilizer equivalent deficit
    fert_input_eq = pmax(0, target_Q_K - Q_K),
    
    # 5. Convert equivalents back to mg/kg K
    fert_input_mgkg = (fert_input_eq / valence["K"]) * molar_mass["K"]
  )

# Display a summary of dynamic I_crit and required fertilizer
summary(fert_recommendations %>% select(I_crit, fert_input_mgkg))
```

## 4 - Parallel Performance Optimization (mclapply)

We utilize `mclapply` from the `parallel` package to distribute a cross-validation of the `nlme` models across the available heavy-compute cores.

```{r}
library(parallel)
cores <- detectCores()
cat("Using", cores, "cores for LOOCV optimization...\n")

# Mock parallel workload fitting models on subsamples
set.seed(42)
folds <- sample(1:4, nrow(yield_data_clean), replace = TRUE)
loocv_results <- mclapply(1:4, function(fold) {
  test_indices <- which(folds == fold)
  
  # Ensure we only use the clean dataset
  sample_data <- yield_data_clean[-test_indices, ]
  
  m_sub <- nlme(
    annual_yield_mp_DM ~ k_model(z_AR_K, z_b, A, Y_0, c_I, c_b),
    data = sample_data,
    fixed = my_fixed,
    random = Y_0 ~ 1 | site/year,
    start = my_starts,
    weights = varPower(form = ~fitted(.)),
    control = nlmeControl(maxIter = 200, pnlsMaxIter = 100)
  )
  return(fixef(m_sub))
}, mc.cores = min(4, cores))

print(loocv_results)
```