---
title: "Bayesian Phosphorus Yield Modeling"
author: "Antigravity Agent"
format: html
---

## Setup and Data Loading

```{r setup, include=FALSE}
library(dplyr)
library(brms)

# Create models directory if it doesn't exist
if (!dir.exists("models")) {
  dir.create("models")
}

# Load the prepared dataset
D_Yield <- readRDS("models/D_Yield.rds")

# Remove rows with NA in the crucial predictors
d_brms <- D_Yield |>
  filter(!is.na(total_yield),
         !is.na(soil_0_20_P_AAE10),
         !is.na(soil_0_20_P_CO2),
         !is.na(a_CO2_total_mg_L),
         !is.na(z_inv_b),
         !is.na(z_inv_b_cons),
         !is.na(z_fert_N),
         !is.na(z_Temp_Mean),
         !is.na(z_Prec_Anom),
         !is.na(crop),
         !is.na(site),
         !is.na(year)) |>
  # Crop factor is necessary for brms group-level effects
  mutate(crop = as.factor(crop),
         site = as.factor(site),
         year = as.factor(year))
```

## Define Priors and Configuration

We use biological bounds to prevent the frequentist singularities (D -> infinity, c -> 0).

```{r priors}
# Common priors for the three Yield models
bprior_yield <- c(
  prior(normal(50, 50), nlpar = "Y0", lb = 0),     
  prior(normal(100, 50), nlpar = "A", lb = 0),     
  prior(lognormal(-2, 2), nlpar = "cbase", lb = 0), 
  prior(normal(0, 1), nlpar = "betainvb"),
  prior(normal(0, 1), nlpar = "betaN"),
  prior(normal(0, 1), nlpar = "betaTemp"),
  prior(normal(0, 1), nlpar = "betaPrec")
)

# Priors for the Uptake model (Michaelis-Menten)
bprior_uptake <- c(
  prior(normal(30, 15), nlpar = "Vmax", lb = 0),
  prior(lognormal(1, 1), nlpar = "Kbase", lb = 0),
  prior(normal(0, 1), nlpar = "betainvb"),
  prior(normal(0, 1), nlpar = "betaN"),
  prior(normal(0, 1), nlpar = "betaTemp"),
  prior(normal(0, 1), nlpar = "betaPrec")
)

# Priors for the CP model (Mitscherlich)
bprior_cp <- c(
  prior(normal(1.5, 1), nlpar = "Cmin", lb = 0),
  prior(normal(2.0, 1.5), nlpar = "DeltaC", lb = 0),
  prior(lognormal(-2, 2), nlpar = "cbase", lb = 0),
  prior(normal(0, 1), nlpar = "betainvb"),
  prior(normal(0, 1), nlpar = "betaTemp"),
  prior(normal(0, 1), nlpar = "betaPrec")
)

# HMC Sampling config
sampling_iter <- 2000
sampling_chains <- 4
sampling_cores <- 4 
```

## Model 1: Empirical Quantity ($P_{\text{AAE10}}$)

```{r fit-aae10}
bform_aae <- bf(
  total_yield ~ Y0 + (A - Y0) * (1 - exp(-(cbase * exp(betainvb * z_inv_b_cons + betaN * z_fert_N + betaTemp * z_Temp_Mean + betaPrec * z_Prec_Anom) * soil_0_20_P_AAE10))),
  Y0 ~ crop - 1 + (1 | site/year),
  A ~ crop - 1 + (1 | site/year),
  cbase ~ crop - 1,
  betainvb + betaN + betaTemp + betaPrec ~ 1,
  nl = TRUE
)

# Fit model (uses cmdstanr or rstan backend natively if configured, we just run brm)
m_brms_raw_aae <- brm(
  formula = bform_aae,
  data = d_brms,
  prior = bprior_yield,
  family = gaussian(),
  backend = "cmdstanr",
  threads = threading(2),
  chains = sampling_chains, 
  cores = sampling_cores, 
  iter = sampling_iter,
  control = list(adapt_delta = 0.95),
  seed = 42,
  file = "models/brms_yield_raw_aae"
)

metrics_file_1 <- "models/brms_yield_raw_aae_metrics.rds"
if (!file.exists(metrics_file_1)) {
  r2 <- bayes_R2(m_brms_raw_aae, re_formula = NA)
  saveRDS(r2, metrics_file_1)
}

rm(m_brms_raw_aae)
gc()
```

## Model 2: Empirical Intensity ($P_{\text{CO}_2}$)

```{r fit-co2}
bform_co2 <- bf(
  total_yield ~ Y0 + (A - Y0) * (1 - exp(-(cbase * exp(betainvb * z_inv_b + betaN * z_fert_N + betaTemp * z_Temp_Mean + betaPrec * z_Prec_Anom) * soil_0_20_P_CO2))),
  Y0 ~ crop - 1 + (1 | site/year),
  A ~ crop - 1 + (1 | site/year),
  cbase ~ crop - 1,
  betainvb + betaN + betaTemp + betaPrec ~ 1,
  nl = TRUE
)

m_brms_raw_co2 <- brm(
  formula = bform_co2,
  data = d_brms,
  prior = bprior_yield,
  family = gaussian(),
  backend = "cmdstanr",
  threads = threading(2),
  chains = sampling_chains, 
  cores = sampling_cores, 
  iter = sampling_iter,
  control = list(adapt_delta = 0.95),
  seed = 42,
  file = "models/brms_yield_raw_co2"
)

metrics_file_2 <- "models/brms_yield_raw_co2_metrics.rds"
if (!file.exists(metrics_file_2)) {
  r2 <- bayes_R2(m_brms_raw_co2, re_formula = NA)
  saveRDS(r2, metrics_file_2)
}

rm(m_brms_raw_co2)
gc()
```

## Model 3: Thermodynamic Activity ($a_{\text{CO}_2}$)

```{r fit-thm}
bform_thm <- bf(
  total_yield ~ Y0 + (A - Y0) * (1 - exp(-(cbase * exp(betainvb * z_inv_b + betaN * z_fert_N + betaTemp * z_Temp_Mean + betaPrec * z_Prec_Anom) * a_CO2_total_mg_L))),
  Y0 ~ crop - 1 + (1 | site/year),
  A ~ crop - 1 + (1 | site/year),
  cbase ~ crop - 1,
  betainvb + betaN + betaTemp + betaPrec ~ 1,
  nl = TRUE
)

m_brms_thm_co2 <- brm(
  formula = bform_thm,
  data = d_brms,
  prior = bprior_yield,
  family = gaussian(),
  backend = "cmdstanr",
  threads = threading(2),
  chains = sampling_chains, 
  cores = sampling_cores, 
  iter = sampling_iter,
  control = list(adapt_delta = 0.95),
  seed = 42,
  file = "models/brms_yield_thm_co2"
)

metrics_file_3 <- "models/brms_yield_thm_co2_metrics.rds"
if (!file.exists(metrics_file_3)) {
  r2 <- bayes_R2(m_brms_thm_co2, re_formula = NA)
  saveRDS(r2, metrics_file_3)
}

rm(m_brms_thm_co2)
gc()
```

## Phosphorus Uptake ($P_{\text{uptake}}$) Models

### Model 4: Uptake vs Empirical Quantity ($P_{\text{AAE10}}$)
```{r fit-uptake-aae}
bform_uptake_aae <- bf(
  annual_P_uptake ~ (Vmax * soil_0_20_P_AAE10) / ((Kbase * exp(betainvb * z_inv_b_cons + betaN * z_fert_N + betaTemp * z_Temp_Mean + betaPrec * z_Prec_Anom)) + soil_0_20_P_AAE10),
  Vmax ~ crop - 1 + (1 | site/year),
  Kbase ~ crop - 1,
  betainvb + betaN + betaTemp + betaPrec ~ 1,
  nl = TRUE
)

m_brms_uptake_aae <- brm(
  formula = bform_uptake_aae,
  data = d_brms,
  prior = bprior_uptake,
  family = gaussian(),
  backend = "cmdstanr",
  threads = threading(2),
  chains = sampling_chains, 
  cores = sampling_cores, 
  iter = sampling_iter,
  control = list(adapt_delta = 0.95),
  seed = 42,
  file = "models/brms_uptake_raw_aae"
)

metrics_file_4 <- "models/brms_uptake_raw_aae_metrics.rds"
if (!file.exists(metrics_file_4)) {
  r2 <- bayes_R2(m_brms_uptake_aae, re_formula = NA)
  saveRDS(r2, metrics_file_4)
}

rm(m_brms_uptake_aae)
gc()
```

### Model 5: Uptake vs Empirical Intensity ($P_{\text{CO}_2}$)
```{r fit-uptake-co2}
bform_uptake_co2 <- bf(
  annual_P_uptake ~ (Vmax * soil_0_20_P_CO2) / ((Kbase * exp(betainvb * z_inv_b_cons + betaN * z_fert_N + betaTemp * z_Temp_Mean + betaPrec * z_Prec_Anom)) + soil_0_20_P_CO2),
  Vmax ~ crop - 1 + (1 | site/year),
  Kbase ~ crop - 1,
  betainvb + betaN + betaTemp + betaPrec ~ 1,
  nl = TRUE
)

m_brms_uptake_co2 <- brm(
  formula = bform_uptake_co2,
  data = d_brms,
  prior = bprior_uptake,
  family = gaussian(),
  backend = "cmdstanr",
  threads = threading(2),
  chains = sampling_chains, 
  cores = sampling_cores, 
  iter = sampling_iter,
  control = list(adapt_delta = 0.95),
  seed = 42,
  file = "models/brms_uptake_raw_co2"
)

metrics_file_5 <- "models/brms_uptake_raw_co2_metrics.rds"
if (!file.exists(metrics_file_5)) {
  r2 <- bayes_R2(m_brms_uptake_co2, re_formula = NA)
  saveRDS(r2, metrics_file_5)
}

rm(m_brms_uptake_co2)
gc()
```

### Model 6: Uptake vs Thermodynamic Activity ($a_{\text{CO}_2}$)
```{r fit-uptake-thm}
bform_uptake_thm <- bf(
  annual_P_uptake ~ (Vmax * a_CO2_total_mg_L) / ((Kbase * exp(betainvb * z_inv_b_cons + betaN * z_fert_N + betaTemp * z_Temp_Mean + betaPrec * z_Prec_Anom)) + a_CO2_total_mg_L),
  Vmax ~ crop - 1 + (1 | site/year),
  Kbase ~ crop - 1,
  betainvb + betaN + betaTemp + betaPrec ~ 1,
  nl = TRUE
)

m_brms_uptake_thm <- brm(
  formula = bform_uptake_thm,
  data = d_brms,
  prior = bprior_uptake,
  family = gaussian(),
  backend = "cmdstanr",
  threads = threading(2),
  chains = sampling_chains, 
  cores = sampling_cores, 
  iter = sampling_iter,
  control = list(adapt_delta = 0.95),
  seed = 42,
  file = "models/brms_uptake_thm_co2"
)

metrics_file_6 <- "models/brms_uptake_thm_co2_metrics.rds"
if (!file.exists(metrics_file_6)) {
  r2 <- bayes_R2(m_brms_uptake_thm, re_formula = NA)
  saveRDS(r2, metrics_file_6)
}

rm(m_brms_uptake_thm)
gc()
```


## Critical P Concentration ($C_P$) Models

```{r setup-cp}
d_brms_cp <- d_brms |> 
  mutate(C_P = annual_P_uptake / annual_yield_mp_DM) |> 
  filter(is.finite(C_P), C_P > 0)
```

### Model 7: CP vs Empirical Quantity ($P_{\text{AAE10}}$)
```{r fit-cp-aae}
bform_cp_aae <- bf(
  C_P ~ Cmin + DeltaC * (1 - exp(-(cbase * exp(betainvb * z_inv_b_cons + betaTemp * z_Temp_Mean + betaPrec * z_Prec_Anom)) * soil_0_20_P_AAE10)),
  Cmin ~ crop - 1 + (1 | site/year),
  DeltaC ~ crop - 1 + (1 | site/year),
  cbase ~ crop - 1,
  betainvb + betaTemp + betaPrec ~ 1,
  nl = TRUE
)

m_brms_cp_aae <- brm(
  formula = bform_cp_aae,
  data = d_brms_cp,
  prior = bprior_cp,
  family = gaussian(),
  backend = "cmdstanr",
  threads = threading(2),
  chains = sampling_chains, 
  cores = sampling_cores, 
  iter = sampling_iter,
  control = list(adapt_delta = 0.95),
  seed = 42,
  file = "models/brms_cp_raw_aae"
)

metrics_file_7 <- "models/brms_cp_raw_aae_metrics.rds"
if (!file.exists(metrics_file_7)) {
  r2 <- bayes_R2(m_brms_cp_aae, re_formula = NA)
  saveRDS(r2, metrics_file_7)
}

rm(m_brms_cp_aae)
gc()
```

### Model 8: CP vs Empirical Intensity ($P_{\text{CO}_2}$)
```{r fit-cp-co2}
bform_cp_co2 <- bf(
  C_P ~ Cmin + DeltaC * (1 - exp(-(cbase * exp(betainvb * z_inv_b_cons + betaTemp * z_Temp_Mean + betaPrec * z_Prec_Anom)) * soil_0_20_P_CO2)),
  Cmin ~ crop - 1 + (1 | site/year),
  DeltaC ~ crop - 1 + (1 | site/year),
  cbase ~ crop - 1,
  betainvb + betaTemp + betaPrec ~ 1,
  nl = TRUE
)

m_brms_cp_co2 <- brm(
  formula = bform_cp_co2,
  data = d_brms_cp,
  prior = bprior_cp,
  family = gaussian(),
  backend = "cmdstanr",
  threads = threading(2),
  chains = sampling_chains, 
  cores = sampling_cores, 
  iter = sampling_iter,
  control = list(adapt_delta = 0.95),
  seed = 42,
  file = "models/brms_cp_raw_co2"
)

metrics_file_8 <- "models/brms_cp_raw_co2_metrics.rds"
if (!file.exists(metrics_file_8)) {
  r2 <- bayes_R2(m_brms_cp_co2, re_formula = NA)
  saveRDS(r2, metrics_file_8)
}

rm(m_brms_cp_co2)
gc()
```

### Model 9: CP vs Thermodynamic Activity ($a_{\text{CO}_2}$)
```{r fit-cp-thm}
bform_cp_thm <- bf(
  C_P ~ Cmin + DeltaC * (1 - exp(-(cbase * exp(betainvb * z_inv_b_cons + betaTemp * z_Temp_Mean + betaPrec * z_Prec_Anom)) * a_CO2_total_mg_L)),
  Cmin ~ crop - 1 + (1 | site/year),
  DeltaC ~ crop - 1 + (1 | site/year),
  cbase ~ crop - 1,
  betainvb + betaTemp + betaPrec ~ 1,
  nl = TRUE
)

m_brms_cp_thm <- brm(
  formula = bform_cp_thm,
  data = d_brms_cp,
  prior = bprior_cp,
  family = gaussian(),
  backend = "cmdstanr",
  threads = threading(2),
  chains = sampling_chains, 
  cores = sampling_cores, 
  iter = sampling_iter,
  control = list(adapt_delta = 0.95),
  seed = 42,
  file = "models/brms_cp_thm_co2"
)

metrics_file_9 <- "models/brms_cp_thm_co2_metrics.rds"
if (!file.exists(metrics_file_9)) {
  r2 <- bayes_R2(m_brms_cp_thm, re_formula = NA)
  saveRDS(r2, metrics_file_9)
}

rm(m_brms_cp_thm)
gc()
```

All 9 models have been sequentially compiled, sampled, cached, and flushed from memory!
