---
title: "Data Sanity Checks"
format: html
editor: visual
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readxl)
library(ggpmisc)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# 1. Load legacy kinetics and climate data
RES <- readRDS("../data/RES.rds")
D_legacy <- RES$D
climate_data <- readRDS("../data/all_P.rds") |>
    dplyr::select(site, year, anavg_temp, ansum_prec, juvdev_temp, juvdev_prec) |>
    dplyr::distinct()

site_geochemistry <- D_legacy |>
    group_by(site) |>
    summarise(feox_mean = mean(Feox, na.rm = TRUE), alox_mean = mean(Alox, na.rm = TRUE)) |>
    ungroup()

kinetics_stable <- D_legacy |>
    dplyr::select(site, treatment_ID, rep, k, v0_kPS = kPS, Pmax_PS = PS) |>
    distinct(site, treatment_ID, rep, .keep_all = TRUE)

# 2. Load accurate raw STYCS dataset and construct D_master
D_master <- read_excel("../data/STYCS_data_2023_260511.xlsx", guess_max = 20000) |>
    rename(rep = replicate) |>
    mutate(site = gsub("STYCS_", "", LtE_name)) |>
    left_join(climate_data, by = c("site", "year")) |>
    mutate(
        soil_0_20_P_CO2 = soil_0_20_P_test * 0.155,
        crop = crop_abr,
        fert_P_tot = fert_P2O5_tot / 2.291,
        annual_P_uptake = rowSums(across(starts_with("P_harv")), na.rm = TRUE),
        annual_yield_mp_DM = rowSums(across(matches("^harv.*mp_yield_DM$")), na.rm = TRUE),
        annual_yield_bp_DM = rowSums(across(matches("^harv.*bp[1-2]_yield_DM$")), na.rm = TRUE),
        annual_P_balance = fert_P_tot - annual_P_uptake
    ) |>
    filter(year >= 1990) |>
    left_join(site_geochemistry, by = "site") |>
    left_join(kinetics_stable, by = c("site", "treatment_ID", "rep")) |>
    # Focus only on the plots where we have stable kinetics (the main P treatments)
    filter(!is.na(Pmax_PS)) |>
    # Just to map Treatment to treatment_ID for plotting aesthetics
    mutate(Treatment = factor(treatment_ID, levels = c("P0", "P100", "P166")))
```

## Introduction

This notebook contains sanity checks for the agronomic data using the **newly constructed `D_master`**, which merges the correct, uncorrupted field measurements from the raw Excel file with the stable kinetic laboratory measurements from the legacy `RES.rds` dataset. 

## Summary Table

Let's look at the average applied P fertilizer (`fert_P_tot`) and yield for each treatment at each site.

```{r}
summary_table <- D_master %>%
    group_by(site, Treatment) %>%
    summarise(
        mean_yield = mean(annual_yield_mp_DM, na.rm = TRUE),
        mean_Fert_P = mean(fert_P_tot, na.rm = TRUE),
        .groups = "drop"
    )

knitr::kable(summary_table, digits = 2)
```

**Observation:** The anomaly is now entirely resolved! For Cadenazzo, `P0` correctly shows the lowest yield and 0 applied fertilizer, whereas `P166` correctly shows the highest yield and highest fertilizer applied.

## Visualizations

### 1. Yield by Treatment and Site

```{r fig.width=10, fig.height=8}
ggplot(D_master, aes(x = Treatment, y = annual_yield_mp_DM, fill = Treatment)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~site, scales = "free_y") +
    labs(
        title = "Annual Dry Matter Yield by Treatment and Site",
        x = "Treatment",
        y = "Annual Yield (kg DM / ha)"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 2. Applied P Fertilizer by Treatment and Site

```{r fig.width=10, fig.height=8}
ggplot(D_master, aes(x = Treatment, y = fert_P_tot, fill = Treatment)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~site, scales = "free_y") +
    labs(
        title = "Total Applied P Fertilizer (fert_P_tot) by Treatment and Site",
        x = "Treatment",
        y = "Applied P Fertilizer (kg P / ha)"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 3. Yield vs. Applied P Fertilizer

```{r fig.width=10, fig.height=8}
ggplot(D_master, aes(x = fert_P_tot, y = annual_yield_mp_DM, color = Treatment)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
    facet_wrap(~site, scales = "free") +
    labs(
        title = "Yield vs. Actual Applied P Fertilizer",
        x = "Total Applied P Fertilizer (fert_P_tot)",
        y = "Annual Yield (kg DM / ha)"
    ) +
    theme_bw()
```

### 4. Soil P vs. Treatment

```{r fig.width=10, fig.height=8}
ggplot(D_master, aes(x = Treatment, y = soil_0_20_P_CO2, fill = Treatment)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~site, scales = "free_y") +
    labs(
        title = "Soil P (CO2 extractable) by Treatment and Site",
        x = "Treatment",
        y = "P_CO2 (mg/kg)"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 5. P Uptake vs. Applied Fertilizer

```{r fig.width=10, fig.height=8}
ggplot(D_master, aes(x = fert_P_tot, y = annual_P_uptake, color = Treatment)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
    facet_wrap(~site, scales = "free") +
    labs(
        title = "Annual P Uptake vs. Applied P Fertilizer",
        x = "Total Applied P Fertilizer (fert_P_tot)",
        y = "Annual P Uptake (kg P / ha)"
    ) +
    theme_bw()
```

### 6. Cumulated P Balance vs. Soil P Indicators

Here we calculate the cumulated `P_balance` across the entire trial duration (1990-2022) for each plot, and compare it against the mean soil test indicators ($P_{CO2}$, $P_{AAE10}$, and $P_{desorb} / PS$).

```{r fig.width=12, fig.height=15}
# Calculate cumulative P balance per plot
D_cum <- D_master %>%
    group_by(site, Treatment, rep) %>%
    summarise(
        Cumulated_P_Balance = sum(annual_P_balance, na.rm = TRUE),
        mean_P_CO2 = mean(soil_0_20_P_CO2, na.rm = TRUE),
        mean_P_AAE10 = mean(soil_0_20_P_AAE10, na.rm = TRUE),
        mean_PS = mean(Pmax_PS, na.rm = TRUE),
        .groups = "drop"
    )

# Pivot the dataset to long format for faceting by indicator
D_indicators <- D_cum %>%
    pivot_longer(
        cols = c(mean_P_CO2, mean_P_AAE10, mean_PS),
        names_to = "Indicator",
        values_to = "Indicator_Value"
    ) %>%
    # Make the facet labels nicer
    mutate(Indicator = recode(Indicator,
        mean_P_CO2 = "P_CO2 (Water Soluble)",
        mean_P_AAE10 = "P_AAE10 (Chelate Extractable)",
        mean_PS = "PS (Desorbable P / Capacity)"))

ggplot(D_indicators, aes(x = Indicator_Value, y = Cumulated_P_Balance, color = site)) +
    geom_point(alpha = 0.8, size = 2) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
    facet_grid(site ~ Indicator, scales = "free") +
    labs(
        title = "Cumulated P Balance (1990-2022) vs. Soil P Indicators",
        x = "Measured Soil P Indicator Value",
        y = "Cumulated P Balance (kg P / ha)",
        color = "Site"
    ) +
    theme_bw()
```

### 7. Annual P Uptake vs. Soil P Indicators

To compare the predictive power of the soil tests against the annual P-uptake, we plot linear regressions faceted by the indicator. We include the linear regression equation and the $R^2$ to assess meaningful differences.

```{r fig.width=12, fig.height=15}
# Pivot the dataset to long format for faceting by indicator
D_uptake_annual <- D_master %>%
    pivot_longer(
        cols = c(soil_0_20_P_CO2, soil_0_20_P_AAE10, Pmax_PS),
        names_to = "Indicator",
        values_to = "Indicator_Value"
    ) %>%
    mutate(Indicator = recode(Indicator,
        soil_0_20_P_CO2 = "P_CO2 (Water Soluble)",
        soil_0_20_P_AAE10 = "P_AAE10 (Chelate Extractable)",
        Pmax_PS = "PS (Desorbable P / Capacity)")) %>%
    # Filter missing to avoid plotting warnings
    filter(!is.na(Indicator_Value) & !is.na(annual_P_uptake))

ggplot(D_uptake_annual, aes(x = Indicator_Value, y = annual_P_uptake)) +
    geom_point(aes(color = site), alpha = 0.6) +
    geom_smooth(method = "lm", color = "black", se = FALSE) +
    stat_poly_eq(use_label(c("eq", "R2"))) +
    facet_grid(site ~ Indicator, scales = "free") +
    labs(
        title = "Annual P Uptake vs. Soil P Indicators",
        x = "Measured Soil P Indicator Value",
        y = "Annual P Uptake (kg P / ha)",
        color = "Site"
    ) +
    theme_bw()
```

### 8. Cumulated P Uptake vs. Soil P Indicators

Here we evaluate the long-term predictive power. We sum the `annual_P_uptake` over the 1990-2022 period for each plot, and regress it against the soil test indicators.

```{r fig.width=12, fig.height=15}
# Calculate cumulative P uptake per plot
D_cum_uptake <- D_master %>%
    group_by(site, Treatment, rep) %>%
    summarise(
        Cumulated_P_Uptake = sum(annual_P_uptake, na.rm = TRUE),
        mean_P_CO2 = mean(soil_0_20_P_CO2, na.rm = TRUE),
        mean_P_AAE10 = mean(soil_0_20_P_AAE10, na.rm = TRUE),
        mean_PS = mean(Pmax_PS, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    # Pivot for faceting
    pivot_longer(
        cols = c(mean_P_CO2, mean_P_AAE10, mean_PS),
        names_to = "Indicator",
        values_to = "Indicator_Value"
    ) %>%
    mutate(Indicator = recode(Indicator,
        mean_P_CO2 = "P_CO2 (Water Soluble)",
        mean_P_AAE10 = "P_AAE10 (Chelate Extractable)",
        mean_PS = "PS (Desorbable P / Capacity)")) %>%
    filter(!is.na(Indicator_Value) & !is.na(Cumulated_P_Uptake))

ggplot(D_cum_uptake, aes(x = Indicator_Value, y = Cumulated_P_Uptake)) +
    geom_point(aes(color = site), alpha = 0.8, size = 2) +
    geom_smooth(method = "lm", color = "black", se = FALSE) +
    stat_poly_eq(use_label(c("eq", "R2")), label.y = 0.95) +
    facet_grid(site ~ Indicator, scales = "free") +
    labs(
        title = "Cumulated P Uptake (1990-2022) vs. Soil P Indicators",
        x = "Measured Soil P Indicator Value",
        y = "Cumulated P Uptake (kg P / ha)",
        color = "Site"
    ) +
    theme_bw()
```
