---
title: "Data Sanity Checks"
format: html
editor: visual
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readxl)
library(ggpmisc)
library(tikzDevice)
options(tikzDefaultEngine = "luatex")
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  dev = "tikz", fig.align = "center"
)

# 1. Load legacy kinetics and climate data
RES <- readRDS("data/RES.rds")
D_legacy <- RES$D
climate_data <- readRDS("data/all_P.rds") |>
    dplyr::select(site, year, anavg_temp, ansum_prec, juvdev_temp, juvdev_prec) |>
    dplyr::distinct()

site_geochemistry <- D_legacy |>
    group_by(site) |>
    summarise(feox_mean = mean(Feox, na.rm = TRUE), alox_mean = mean(Alox, na.rm = TRUE)) |>
    ungroup()

kinetics_stable <- D_legacy |>
    dplyr::select(site, treatment_ID, rep, k, v0_kPS = kPS, Pmax_PS = PS) |>
    distinct(site, treatment_ID, rep, .keep_all = TRUE)

# 2. Load accurate raw STYCS dataset and construct D_main
D_main <- read_excel("data/STYCS_data_2023_260511.xlsx", guess_max = 20000) |>
    rename(rep = replicate) |>
    mutate(site = gsub("STYCS_", "", LtE_name)) |>
    left_join(climate_data, by = c("site", "year")) |>
    mutate(
        soil_0_20_P_CO2 = soil_0_20_P_test * 0.155,
        crop = crop_abr,
        fert_P_tot = fert_P2O5_tot / 2.291,
        annual_P_uptake = rowSums(across(starts_with("P_harv")), na.rm = TRUE),
        annual_yield_mp_DM = rowSums(across(matches("^harv.*mp_yield_DM$")), na.rm = TRUE),
        annual_yield_bp_DM = rowSums(across(matches("^harv.*bp[1-2]_yield_DM$")), na.rm = TRUE),
        annual_P_balance = fert_P_tot - annual_P_uptake
    ) |>
    filter(year >= 1990) |>
    left_join(site_geochemistry, by = "site") |>
    left_join(kinetics_stable, by = c("site", "treatment_ID", "rep")) |>
    # Focus only on the plots where we have stable kinetics (the main P treatments)
    filter(!is.na(Pmax_PS)) |>
    # Just to map Treatment to treatment_ID for plotting aesthetics
    mutate(Treatment = factor(treatment_ID, levels = c("P0", "P100", "P166")))
```

## Introduction

This notebook contains sanity checks for the agronomic data using the **newly constructed `D_main`**, which merges the correct, uncorrupted field measurements from the raw Excel file with the stable kinetic laboratory measurements from the legacy `RES.rds` dataset.

## Summary Table

Let's look at the average applied P fertilizer (`fert_P_tot`) and yield for each treatment at each site.

```{r}
summary_table <- D_main %>%
    group_by(site, Treatment) %>%
    summarise(
        mean_yield = mean(annual_yield_mp_DM, na.rm = TRUE),
        mean_Fert_P = mean(fert_P_tot, na.rm = TRUE),
        .groups = "drop"
    )

knitr::kable(summary_table, digits = 2)
```

**Observation:** The anomaly is now entirely resolved! For Cadenazzo, `P0` correctly shows the lowest yield and 0 applied fertilizer, whereas `P166` correctly shows the highest yield and highest fertilizer applied.

## Visualizations

### 1. Yield by Treatment and Site

```{r fig.width=10, fig.height=8}
ggplot(D_main, aes(x = Treatment, y = annual_yield_mp_DM, fill = Treatment)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~site, scales = "free_y") +
    labs(
        title = "Annual Dry Matter Yield by Treatment \\& Site",
        x = "Treatment",
        y = "Annual Yield (kg DM / ha)"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 2. Applied P Fertilizer by Treatment and Site

```{r fig.width=10, fig.height=8}
ggplot(D_main, aes(x = Treatment, y = fert_P_tot, fill = Treatment)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~site, scales = "free_y") +
    labs(
        title = "Total Applied P Fertilizer by Treatment \\& Site",
        x = "Treatment",
        y = "Applied P Fertilizer (kg P / ha)"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 3. Yield vs. Applied P Fertilizer

```{r fig.width=10, fig.height=8}
ggplot(D_main, aes(x = fert_P_tot, y = annual_yield_mp_DM, color = Treatment)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
    facet_wrap(~site, scales = "free") +
    labs(
        title = "Yield vs. Actual Applied P Fertilizer",
        x = "Total Applied P Fertilizer",
        y = "Annual Yield (kg DM / ha)"
    ) +
    theme_bw()
```

### 4. Soil P vs. Treatment

```{r fig.width=10, fig.height=8}
ggplot(D_main, aes(x = Treatment, y = soil_0_20_P_CO2, fill = Treatment)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~site, scales = "free_y") +
    labs(
        title = "Soil $P$ ($CO_2$ extractable) by Treatment \\& Site",
        x = "Treatment",
        y = "$P_{CO_2}$ (mg/kg)"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 5. P Uptake vs. Applied Fertilizer

```{r fig.width=10, fig.height=8}
ggplot(D_main, aes(x = fert_P_tot, y = annual_P_uptake, color = Treatment)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
    facet_wrap(~site, scales = "free") +
    labs(
        title = "Annual P Uptake vs. Applied P Fertilizer",
        x = "Total Applied P Fertilizer",
        y = "Annual P Uptake (kg P / ha)"
    ) +
    theme_bw()
```

### 6. Cumulated P Balance vs. Soil P Indicators

Here we calculate the cumulated `P_balance` across the entire trial duration (1990-2022) for each plot, and compare it against the mean soil test indicators ($P_{CO2}$, $P_{AAE10}$, and $P_{desorb} / PS$).

```{r fig.width=12, fig.height=15}
# Calculate cumulative P balance per plot
D_cum <- D_main %>%
    group_by(site, Treatment, rep) %>%
    summarise(
        Cumulated_P_Balance = sum(annual_P_balance, na.rm = TRUE),
        mean_P_CO2 = mean(soil_0_20_P_CO2, na.rm = TRUE),
        mean_P_AAE10 = mean(soil_0_20_P_AAE10, na.rm = TRUE),
        mean_PS = mean(Pmax_PS, na.rm = TRUE),
        v0_kPS = mean(v0_kPS, na.rm = TRUE),
        mean_Clay = mean(soil_0_20_clay, na.rm = TRUE),
        mean_Silt = mean(soil_0_20_silt, na.rm = TRUE),
        mean_pH = mean(soil_0_20_pH_H2O, na.rm = TRUE),
        mean_Corg = mean(soil_0_20_Corg, na.rm = TRUE),
        mean_Prec = mean(ansum_prec, na.rm = TRUE),
        mean_Temp = mean(anavg_temp, na.rm = TRUE),
        mean_K = mean(soil_0_20_K_AAE10, na.rm = TRUE),
        mean_Mg = mean(soil_0_20_Mg_AAE10, na.rm = TRUE),
        mean_Lime = mean(soil_0_20_presample_lime, na.rm = TRUE),
        .groups = "drop"
    )

# Pivot the dataset to long format for faceting by indicator
D_indicators <- D_cum %>%
    pivot_longer(
        cols = c(mean_P_CO2, mean_P_AAE10, mean_PS),
        names_to = "Indicator",
        values_to = "Indicator_Value"
    ) %>%
    # Make the facet labels nicer
    mutate(Indicator = recode(Indicator,
        mean_P_CO2 = "P_CO2 (Water Soluble)",
        mean_P_AAE10 = "P_AAE10 (Chelate Extractable)",
        mean_PS = "PS (Desorbable P / Capacity)"))

ggplot(D_indicators, aes(x = Indicator_Value, y = Cumulated_P_Balance, color = site)) +
    geom_point(alpha = 0.8, size = 2) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
    facet_grid(site ~ Indicator, scales = "free") +
    labs(
        title = "Cumulated P Balance (1990--2022) vs. Soil P Indicators",
        x = "Measured Soil P Indicator Value",
        y = "Cumulated P Balance (kg P / ha)",
        color = "Site"
    ) +
    theme_bw()
```

### 6.5 Cumulated P Balance vs. Pedo-Climatic Site Drivers

Following the variance decomposition (ICC), we want to map the cumulative P balance directly against the highly site-distinguishing variables.

```{r fig.width=14, fig.height=10}
D_pedo <- D_cum %>%
    pivot_longer(
        cols = c(mean_Clay, mean_Silt, mean_pH, mean_Corg, mean_Prec, mean_Temp, mean_K, mean_Mg, mean_Lime),
        names_to = "Pedo_Driver",
        values_to = "Driver_Value"
    )

ggplot(D_pedo, aes(x = Driver_Value, y = Cumulated_P_Balance, color = Treatment)) +
    geom_point(alpha = 0.8, size = 2.5) +
    facet_wrap(~ Pedo_Driver, scales = "free_x", ncol = 3) +
    labs(
        title = "Cumulated P Balance (1990--2022) Driven by Pedoclimatic Traits",
        x = "Pedoclimatic Driver Value (Site Average)",
        y = "Cumulated P Balance (kg P / ha)",
        color = "P Treatment"
    ) +
    theme_bw() +
    theme(legend.position = "bottom")
```

### 7. Annual P Uptake vs. Soil P Indicators

To compare the predictive power of the soil tests against the annual P-uptake, we plot linear regressions faceted by the indicator. We include the linear regression equation and the $R^2$ to assess meaningful differences.

```{r fig.width=12, fig.height=15}
# Pivot the dataset to long format for faceting by indicator
D_uptake_annual <- D_main %>%
    pivot_longer(
        cols = c(soil_0_20_P_CO2, soil_0_20_P_AAE10, Pmax_PS),
        names_to = "Indicator",
        values_to = "Indicator_Value"
    ) %>%
    mutate(Indicator = recode(Indicator,
        soil_0_20_P_CO2 = "P_CO2 (Water Soluble)",
        soil_0_20_P_AAE10 = "P_AAE10 (Chelate Extractable)",
        Pmax_PS = "PS (Desorbable P / Capacity)")) %>%
    # Filter missing to avoid plotting warnings
    filter(!is.na(Indicator_Value) & !is.na(annual_P_uptake))

ggplot(D_uptake_annual, aes(x = Indicator_Value, y = annual_P_uptake)) +
    geom_point(aes(color = site), alpha = 0.6) +
    geom_smooth(method = "lm", color = "black", se = FALSE) +
    stat_poly_eq(use_label(c("eq", "R2"))) +
    facet_grid(site ~ Indicator, scales = "free") +
    labs(
        title = "Annual P Uptake vs. Soil P Indicators",
        x = "Measured Soil P Indicator Value",
        y = "Annual P Uptake (kg P / ha)",
        color = "Site"
    ) +
    theme_bw()
```

### 8. Cumulated P Uptake vs. Soil P Indicators

Here we evaluate the long-term predictive power. We sum the `annual_P_uptake` over the 1990-2022 period for each plot, and regress it against the soil test indicators.

```{r fig.width=12, fig.height=15}
# Calculate cumulative P uptake per plot
D_cum_uptake <- D_main %>%
    group_by(site, Treatment, rep) %>%
    summarise(
        Cumulated_P_Uptake = sum(annual_P_uptake, na.rm = TRUE),
        mean_P_CO2 = mean(soil_0_20_P_CO2, na.rm = TRUE),
        mean_P_AAE10 = mean(soil_0_20_P_AAE10, na.rm = TRUE),
        mean_PS = mean(Pmax_PS, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    # Pivot for faceting
    pivot_longer(
        cols = c(mean_P_CO2, mean_P_AAE10, mean_PS),
        names_to = "Indicator",
        values_to = "Indicator_Value"
    ) %>%
    mutate(Indicator = recode(Indicator,
        mean_P_CO2 = "P_CO2 (Water Soluble)",
        mean_P_AAE10 = "P_AAE10 (Chelate Extractable)",
        mean_PS = "PS (Desorbable P / Capacity)")) %>%
    filter(!is.na(Indicator_Value) & !is.na(Cumulated_P_Uptake))

ggplot(D_cum_uptake, aes(x = Indicator_Value, y = Cumulated_P_Uptake)) +
    geom_point(aes(color = site), alpha = 0.8, size = 2) +
    geom_smooth(method = "lm", color = "black", se = FALSE) +
    stat_poly_eq(use_label(c("eq", "R2")), label.y = 0.95) +
    facet_grid(site ~ Indicator, scales = "free") +
    labs(
        title = "Cumulated P Uptake (1990-2022) vs. Soil P Indicators",
        x = "Measured Soil P Indicator Value",
        y = "Cumulated P Uptake (kg P / ha)",
        color = "Site"
    ) +
    theme_bw()
```

## 9. Site-Dependency of Environmental Covariates (Variance Decomposition)

As observed above, metrics like the cumulative P balance and yield responses appear highly site-dependent. To formally quantify how strictly a covariate is grouped by its site (e.g. soil geology) versus how much it varies independently across plots and years (e.g. fertilizer, weather), we can compute the **Intra-Class Correlation Coefficient (ICC)**.

### The Mixed-Effects Variance Decomposition

To formally quantify how strictly a covariate is grouped by its site (e.g. soil geology) versus how much it varies independently across plots and years (e.g. fertilizer, weather), we compute the **Intra-Class Correlation Coefficient (ICC)**.

For every numeric parameter in the dataset, we fit a random intercept mixed-effects model with no fixed effects (`lmer(scale(variable) ~ 1 + (1 | site))`). This forces the model to decompose the total variance of the parameter into exactly two variance fractions: 1. **Between-Site Variance (**$\sigma^2_{site}$): Variance driven purely by moving from one geographical location to another (geology and macro-climate). 2. **Residual Variance (**$\sigma^2_{residual}$): "Noise" happening *within* the site—such as plot-to-plot differences (caused by P, K, Mg fertilizer treatments) and year-to-year differences (weather anomalies, crop rotations).

The ICC is then calculated as the ratio of site-driven variance to total variance: $$ ICC = \frac{\sigma^2_{site}}{\sigma^2_{site} + \sigma^2_{residual}} $$

- **ICC \> 0.80**: The variable is almost entirely dictated by the geographical site (e.g. `pH` or soil texture). These parameters act as excellent, stable candidates for modeling the physical pedoclimatic baseline (the $1/b$ physical highway).
- **ICC \< 0.30**: The variable fluctuates completely independently of the site. Its variance is heavily driven by local experimental treatments (e.g., `fert_N`, `P_CO2`) or local weather (`Temp_Anom`). Nesting these into the models (like the Mitscherlich rate constant $c_{eff}$) helps capture critical plot-level nutritional variance that the rigid site-level variables are blind to.

```{r site-icc, fig.width=8, fig.height=6}
library(lme4)
library(performance)
library(purrr)

# Define the variables we want to decompose
# We select all numeric variables from D_main, excluding kinetics and identifiers
num_cols <- names(D_main)[sapply(D_main, is.numeric)]
covariates <- setdiff(num_cols, c(
    "year", "rep", "k", "Pmax_PS", "v0_kPS", "soil_0_20_P_test"
))

# Helper function to compute ICC for a single variable
get_icc <- function(var_name, data) {
    # We use a tryCatch just in case a variable is missing or singular
    res <- tryCatch({
        # Drop NAs for this specific variable to fit the lmer
        d_clean <- data |> filter(!is.na(.data[[var_name]]))
        
        # Build formula explicitly because lmer doesn't support the dplyr .data pronoun
        fmla <- as.formula(paste0("scale(`", var_name, "`) ~ 1 + (1 | site)"))
        mod <- lmer(fmla, data = d_clean, control = lmerControl(calc.derivs = FALSE))
        
        # Explicitly extract variance components
        vc <- as.data.frame(VarCorr(mod))
        var_site <- vc$vcov[vc$grp == "site"]
        var_resid <- vc$vcov[vc$grp == "Residual"]
        
        icc_val <- var_site / (var_site + var_resid)
        return(icc_val)
    }, error = function(e) return(NA))
    
    return(res)
}

# Calculate ICC for all covariates
icc_results <- data.frame(
    Variable = covariates,
    ICC = sapply(covariates, get_icc, data = D_main)
) |> 
    filter(!is.na(ICC)) |> # Drop variables that failed to compute
    arrange(desc(ICC)) |>
    mutate(
        Category = case_when(
            ICC > 0.8 ~ "Highly Site-Dependent (Geology/Climate Baseline)",
            ICC > 0.3 ~ "Mixed (Site + Plot/Temporal Effects)",
            TRUE ~ "Highly Local/Temporal (Treatments/Weather)"
        )
    )

# Print nice table
knitr::kable(icc_results, digits = 3, caption = "**Table: Intra-Class Correlation (ICC) by Site.** Quantifies the proportion of variance driven strictly by pedo-climatic site differences.")

# Visualize
ggplot(icc_results, aes(x = reorder(Variable, ICC), y = ICC, fill = Category)) +
    geom_col() +
    coord_flip() +
    scale_y_continuous(limits = c(0, 1.05), breaks = seq(0, 1, 0.2)) +
    labs(
        title = "Variance Explained by Site (Intra-Class Correlation)",
        subtitle = "1.0 = Purely site-driven | 0.0 = Purely plot/year-driven",
        x = "Covariate",
        y = "Intra-Class Correlation (ICC)"
    ) +
    theme_bw(base_size = 12) +
    theme(legend.position = "bottom", legend.direction = "vertical")
```

### 10. Covariate Permutation: Identifying the Best Pedoclimatic Drivers

To rigorously test which combination of pedoclimatic traits best explains the variance across our 7 sites, we fit a series of Mixed-Effects models. For each of our 3 main soil indicators ($P_{CO2}$, $P_{AAE10}$, $PS$), we fit a baseline model: `annual_P_balance ~ scale(Indicator) + (1 | site)`

Then, we systematically loop through every 2-variable combination of our 6 pedoclimatic drivers (`pH`, `Corg`, `Clay`, `Lime`, `Temp`, `Prec`) and fit: `annual_P_balance ~ scale(Indicator) + scale(Var1) + scale(Var2) + (1 | site)`

If the two physical variables successfully explain the site-to-site differences, the Random Intercept variance (`1 | site`) should shrink substantially towards 0.

```{r covariate-permutation, warning=FALSE, message=FALSE}
library(tidyr)
library(lme4)
library(dplyr)
library(performance)

# Define variables for Annual data (D_main)
cova_ann <- c("ansum_prec", "anavg_temp", "soil_0_20_pH_H2O", "soil_0_20_Corg", "soil_0_20_clay", "soil_0_20_presample_lime")
inds_ann <- c("soil_0_20_P_CO2", "soil_0_20_P_AAE10", "Pmax_PS", "v0_kPS")

# Define variables for Cumulated data (D_cum)
cova_cum <- c("mean_Prec", "mean_Temp", "mean_pH", "mean_Corg", "mean_Clay", "mean_Lime")
inds_cum <- c("mean_P_CO2", "mean_P_AAE10", "mean_PS", "v0_kPS")

# Function to run permutations
run_perms <- function(target, inds, covs, dat) {
    d_clean <- dat |> drop_na(all_of(c(target, "site", covs, inds)))
    combos <- combn(covs, 2, simplify = FALSE)
    results <- list()
    for(ind in inds) {
        # Baseline model
        f_base <- as.formula(paste0("scale(", target, ") ~ scale(`", ind, "`) + (1 | site)"))
        mod_base <- lmer(f_base, data = d_clean, control = lmerControl(calc.derivs = FALSE))
        base_vc <- as.data.frame(VarCorr(mod_base))
        base_site_var <- base_vc$vcov[base_vc$grp == "site"]
        
        for(cb in combos) {
            v1 <- cb[1]; v2 <- cb[2]
            f_mod <- as.formula(paste0("scale(", target, ") ~ scale(`", ind, "`) + scale(`", v1, "`) + scale(`", v2, "`) + (1 | site)"))
            mod <- suppressWarnings(tryCatch({
                lmer(f_mod, data = d_clean, control = lmerControl(calc.derivs = FALSE))
            }, error = function(e) NULL))
            
            if(is.null(mod)) next
            
            vc <- as.data.frame(VarCorr(mod))
            site_var <- vc$vcov[vc$grp == "site"]
            
            # Safely calculate Marginal R2
            r2m <- suppressWarnings(tryCatch(as.numeric(performance::r2(mod)$R2_marginal), error=function(e) NA))
            
            results[[length(results)+1]] <- data.frame(
                Soil_Indicator = ind,
                Pedo_Var_1 = v1,
                Pedo_Var_2 = v2,
                R2_Marginal = r2m,
                AIC = AIC(mod),
                Site_Var_Reduction_Pct = (base_site_var - site_var) / base_site_var * 100
            )
        }
    }
    
    bind_rows(results) |> 
        mutate(Stability = ifelse(Site_Var_Reduction_Pct < 0, "Unstable (Collinear)", "Stable")) |>
        filter(Stability == "Stable") |>
        mutate(
            R2_Marginal = round(R2_Marginal, 3),
            Site_Var_Reduction_Pct = round(Site_Var_Reduction_Pct, 1),
            AIC = round(AIC, 1)
        ) |>
        arrange(desc(R2_Marginal)) |>
        select(-Stability)
}

# Run for Annual P-Balance
res_ann <- run_perms("annual_P_balance", inds_ann, cova_ann, D_main)
knitr::kable(head(res_ann, 10), caption = "**Top 10 Stable Models: ANNUAL P-Balance (Sorted by Marginal $R^2$)**")

# Run for Annual Yield
res_yield_ann <- run_perms("annual_yield_mp_DM", inds_ann, cova_ann, D_main)
knitr::kable(head(res_yield_ann, 10), caption = "**Top 10 Stable Models: ANNUAL Yield (Sorted by Marginal $R^2$)**")

# Run for Annual P-Uptake
res_up_ann <- run_perms("annual_P_uptake", inds_ann, cova_ann, D_main)
knitr::kable(head(res_up_ann, 10), caption = "**Top 10 Stable Models: ANNUAL P-Uptake (Sorted by Marginal $R^2$)**")

# Run for Cumulated P-Balance
# Ensure D_cum exists in the environment identically to how we generated it
D_cum <- D_main |>
    group_by(site, Treatment, rep) |>
    summarise(
        Cumulated_P_Balance = sum(annual_P_balance, na.rm = TRUE),
        Cumulated_Yield = sum(annual_yield_mp_DM, na.rm = TRUE),
        Cumulated_P_Uptake = sum(annual_P_uptake, na.rm = TRUE),
        mean_P_CO2 = mean(soil_0_20_P_CO2, na.rm = TRUE),
        mean_P_AAE10 = mean(soil_0_20_P_AAE10, na.rm = TRUE),
        mean_PS = mean(Pmax_PS, na.rm = TRUE),
        v0_kPS = mean(v0_kPS, na.rm = TRUE),
        mean_Clay = mean(soil_0_20_clay, na.rm = TRUE),
        mean_Silt = mean(soil_0_20_silt, na.rm = TRUE),
        mean_pH = mean(soil_0_20_pH_H2O, na.rm = TRUE),
        mean_Corg = mean(soil_0_20_Corg, na.rm = TRUE),
        mean_Prec = mean(ansum_prec, na.rm = TRUE),
        mean_Temp = mean(anavg_temp, na.rm = TRUE),
        mean_K = mean(soil_0_20_K_AAE10, na.rm = TRUE),
        mean_Mg = mean(soil_0_20_Mg_AAE10, na.rm = TRUE),
        mean_Lime = mean(soil_0_20_presample_lime, na.rm = TRUE),
        .groups = "drop"
    )

res_cum <- run_perms("Cumulated_P_Balance", inds_cum, cova_cum, D_cum)
knitr::kable(head(res_cum, 10), caption = "**Top 10 Stable Models: CUMULATED P-Balance (Sorted by Marginal $R^2$)**")

# Run for Cumulated Yield
res_yield_cum <- run_perms("Cumulated_Yield", inds_cum, cova_cum, D_cum)
knitr::kable(head(res_yield_cum, 10), caption = "**Top 10 Stable Models: CUMULATED Yield (Sorted by Marginal $R^2$)**")

# Run for Cumulated P-Uptake
res_up_cum <- run_perms("Cumulated_P_Uptake", inds_cum, cova_cum, D_cum)
knitr::kable(head(res_up_cum, 10), caption = "**Top 10 Stable Models: CUMULATED P-Uptake (Sorted by Marginal $R^2$)**")
```

## 11. Final Task Sanity Graphs: Yield and Uptake Distributions

To verify the overall distributions and responses of Yield and P Uptake across the sites, treatments, years, and specific fertilizers, we plot the requested sanity checks below.

```{r final-sanity-graphs, fig.width=12, fig.height=8, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

# Calculate cumulative P uptake
D_graphs <- D_main |>
    group_by(site, plot_nr) |>
    arrange(year) |>
    mutate(
        cumulative_P_up = cumsum(replace_na(annual_P_uptake, 0))
    ) |>
    ungroup() |>
    filter(!is.na(annual_yield_mp_DM))

# Helper to plot across the 3 metrics
plot_sanity <- function(metric_y, title_y) {
    # 1. Y ~ Treatment faceted by trial
    p1 <- ggplot(D_graphs, aes(x = Treatment, y = .data[[metric_y]], fill = Treatment)) +
        geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
        facet_wrap(~ site, scales = "free_y") +
        theme_bw() + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(title = paste(title_y, "~ Treatment by Site"), x = "Treatment", y = title_y)
    
    # 2. Y ~ year faceted by crop and site
    p2 <- ggplot(D_graphs, aes(x = year, y = .data[[metric_y]], color = crop)) +
        geom_point(alpha = 0.5) +
        facet_grid(crop ~ site, scales = "free_y") +
        theme_bw() +
        theme(strip.text.y = element_text(angle = 0), legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(title = paste(title_y, "~ Year faceted by Crop and Site"), x = "Year", y = title_y)
    
    # 3. Y ~ fert_X for X in {P, K, Mg, N}
    D_fert <- D_graphs |>
        select(site, Treatment, all_of(metric_y), fert_P_tot, fert_K_tot, fert_Mg_tot, fert_N_tot) |>
        pivot_longer(cols = starts_with("fert_"), names_to = "Fertilizer", values_to = "Amount") |>
        mutate(Fertilizer = factor(Fertilizer, levels = c("fert_P_tot", "fert_K_tot", "fert_Mg_tot", "fert_N_tot"),
                                   labels = c("P Applied", "K Applied", "Mg Applied", "N Applied")))
        
    p3 <- ggplot(D_fert, aes(x = Amount, y = .data[[metric_y]], color = site)) +
        geom_point(alpha = 0.3) +
        geom_smooth(method = "lm", se = FALSE) +
        facet_wrap(~ Fertilizer, scales = "free_x") +
        theme_bw() +
        labs(title = paste(title_y, "~ Fertilizer Applied"), x = "Fertilizer Amount (kg/ha)", y = title_y)
    
    print(p1)
    print(p2)
    print(p3)
}

plot_sanity("annual_yield_mp_DM", "Annual Dry Yield (kg/ha)")
plot_sanity("annual_P_uptake", "Annual P Uptake (kg P/ha)")
plot_sanity("cumulative_P_up", "Cumulative P Uptake (kg P/ha)")
```