---
title: "The GRUD Step Functions"
subtitle: "Empirical Classification vs. Mechanistic Reality"
author: "Marc Jerónimo Pérez y Ropero"
date: "2026-08-25"
format:
  revealjs:
    theme: default
    transition: slide
    slide-number: true
    width: 1280
    height: 720
---

## $P_{CO2}$ Model { .scrollable .smaller }

```{r}
#| echo: false
#| message: false
#| warning: false
library(plotly)
library(htmltools)

x_clay <- c("<10", "10-20", "20-30", "30-40", ">40")
y_co2 <- c("0.15", "0.46", "0.77", "1.08", "1.39", "1.70", "2.01", "2.32", "2.63", "2.94", "3.25", "3.56", "3.88", "4.19", "4.50", "≥4.65")
z_co2 <- matrix(c(
  1.5, 1.5, 1.5, 1.4, 1.2,
  1.4, 1.4, 1.3, 1.2, 1.1,
  1.2, 1.2, 1.1, 1.0, 1.0,
  1.0, 1.0, 1.0, 1.0, 0.8,
  1.0, 1.0, 1.0, 0.8, 0.6,
  1.0, 1.0, 0.8, 0.6, 0.0,
  1.0, 0.8, 0.6, 0.0, 0.0,
  0.8, 0.8, 0.4, 0.0, 0.0,
  0.8, 0.6, 0.0, 0.0, 0.0,
  0.6, 0.4, 0.0, 0.0, 0.0,
  0.6, 0.4, 0.0, 0.0, 0.0,
  0.4, 0.0, 0.0, 0.0, 0.0,
  0.4, 0.0, 0.0, 0.0, 0.0,
  0.4, 0.0, 0.0, 0.0, 0.0,
  0.0, 0.0, 0.0, 0.0, 0.0,
  0.0, 0.0, 0.0, 0.0, 0.0
), nrow=16, ncol=5, byrow=TRUE)

# Calculate linear model to prove affinity
df_co2 <- expand.grid(x = 1:5, y = 1:16)
df_co2$z <- as.vector(t(z_co2))
lm_co2 <- lm(z ~ x + y, data = df_co2)
r2_co2 <- round(summary(lm_co2)$r.squared, 2)
```

The GRUD attempted to directly model and correct P-deficits empirically, without explicitly considering the underlying mechanics. It assumes that correcting a P-deficit requires a simple linear multiplier $v$, mapping continuous soil variables into rigid, discrete steps:
$$ v: \mathbb{R}^2 \to \mathbb{Q} $$

**Affinity Test:** We quantified how "affine" this matrix is by fitting a generic linear map $L$:
$$ L(P_{CO2}, \text{clay}) = \alpha \cdot P_{CO2} + \beta \cdot \text{clay} + \gamma $$

The discrete table is essentially a perfect affine plane ($R^2 = `r r2_co2`$).

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8

# Reverse y so 0 is at bottom
z_co2_rev <- z_co2[16:1, ]
y_co2_rev <- rev(y_co2)

# Custom color scale based on GRUD classes
colorscale <- list(
  c(0, "#1f77b4"),   # E (0.0)
  c(0.2, "#1f77b4"),
  c(0.2, "#17becf"), # D (0.4-0.8)
  c(0.6, "#17becf"),
  c(0.6, "#2ca02c"), # C (1.0-1.1)
  c(0.75, "#2ca02c"),
  c(0.75, "#ff7f0e"),# B (1.2-1.4)
  c(0.9, "#ff7f0e"),
  c(0.9, "#d62728"), # A (1.5)
  c(1, "#d62728")
)

annotations_list <- list()
for(i in 1:16) {
  for(j in 1:5) {
    annotations_list[[length(annotations_list) + 1]] <- list(
      x = x_clay[j], y = y_co2_rev[i],
      text = as.character(z_co2_rev[i,j]),
      showarrow = FALSE,
      font = list(color = ifelse(z_co2_rev[i,j] < 0.5, "white", "black"))
    )
  }
}

p1 <- plot_ly(x = x_clay, y = y_co2_rev, z = z_co2_rev, type = "heatmap", colorscale = colorscale, zmin=0, zmax=1.5, showscale=FALSE, height=500) %>%
  layout(title = "P_CO2 Flat Heatmap", xaxis = list(title = "Clay %"), yaxis = list(title = "P_CO2 (mg/kg)"), annotations = annotations_list)

# For 3D surface we need numeric axes for smooth rotating
p2 <- plot_ly(x = 1:5, y = 1:16, z = z_co2, type = "surface", colorscale = colorscale, cmin=0, cmax=1.5, showscale=FALSE, height=500) %>%
  layout(title = "P_CO2 3D Staircase", scene = list(
    xaxis = list(title = "Clay %", tickvals=1:5, ticktext=x_clay),
    yaxis = list(title = "P_CO2 (mg/kg)", tickvals=1:16, ticktext=y_co2),
    zaxis = list(title = "Correction Factor C")
  ))

browsable(tagList(
  div(style="display:flex; justify-content:space-between;",
      div(style="width:48%;", p1),
      div(style="width:48%;", p2)
  )
))
```

---

## $P_{AAE10}$ Model { .scrollable .smaller }

```{r}
#| echo: false
#| message: false
#| warning: false

y_aae10 <- c("0-4.9", "5.0-9.9", "10.0-14.9", "15.0-19.9", "20.0-24.9", "25.0-29.9", "30.0-34.9", "35.0-39.9", "40.0-44.9", "45.0-49.9", "50.0-54.9", "55.0-59.9", "60.0-64.9", "65.0-69.9", "70.0-74.9", "75.0-79.9", "80.0-84.9", "85.0-89.9", "90.0-94.9", "95.0-99.9", "100.0-104.9", "105.0-109.9", "110.0-114.9", "115.0-119.9", "120.0-124.9", "≥ 125.0")
z_aae10 <- matrix(c(
  1.5, 1.5, 1.5, 1.4, 1.4,
  1.5, 1.5, 1.4, 1.4, 1.2,
  1.5, 1.4, 1.4, 1.2, 1.2,
  1.4, 1.4, 1.2, 1.2, 1.0,
  1.4, 1.2, 1.2, 1.0, 1.0,
  1.2, 1.2, 1.2, 1.0, 1.0,
  1.2, 1.0, 1.0, 1.0, 1.0,
  1.2, 1.0, 1.0, 1.0, 1.0,
  1.0, 1.0, 1.0, 1.0, 1.0,
  1.0, 1.0, 1.0, 1.0, 1.0,
  1.0, 1.0, 1.0, 0.8, 0.8,
  1.0, 1.0, 0.8, 0.8, 0.6,
  1.0, 1.0, 0.8, 0.8, 0.6,
  1.0, 0.8, 0.8, 0.6, 0.6,
  0.8, 0.8, 0.8, 0.6, 0.6,
  0.8, 0.8, 0.6, 0.6, 0.4,
  0.8, 0.6, 0.6, 0.4, 0.4,
  0.6, 0.6, 0.6, 0.4, 0.4,
  0.6, 0.6, 0.4, 0.4, 0.0,
  0.6, 0.4, 0.4, 0.0, 0.0,
  0.4, 0.4, 0.4, 0.0, 0.0,
  0.4, 0.4, 0.0, 0.0, 0.0,
  0.4, 0.0, 0.0, 0.0, 0.0,
  0.0, 0.0, 0.0, 0.0, 0.0,
  0.0, 0.0, 0.0, 0.0, 0.0,
  0.0, 0.0, 0.0, 0.0, 0.0
), nrow=26, ncol=5, byrow=TRUE)

# Calculate linear model to prove affinity
df_aae <- expand.grid(x = 1:5, y = 1:26)
df_aae$z <- as.vector(t(z_aae10))
lm_aae <- lm(z ~ x + y, data = df_aae)
r2_aae <- round(summary(lm_aae)$r.squared, 2)
```

When looking at the legacy $P_{AAE10}$ extraction method, the GRUD provides a nearly identical mapping structure.

**Affinity Test:** A linear fit for this matrix confirms it is an almost perfectly smooth ramp ($R^2 = `r r2_aae`$).

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

z_aae_rev <- z_aae10[26:1, ]
y_aae_rev <- rev(y_aae10)

annotations_aae <- list()
for(i in 1:26) {
  for(j in 1:5) {
    annotations_aae[[length(annotations_aae) + 1]] <- list(
      x = x_clay[j], y = y_aae_rev[i],
      text = as.character(z_aae_rev[i,j]),
      showarrow = FALSE,
      font = list(color = ifelse(z_aae_rev[i,j] < 0.5, "white", "black"))
    )
  }
}

p3 <- plot_ly(x = x_clay, y = y_aae_rev, z = z_aae_rev, type = "heatmap", colorscale = colorscale, zmin=0, zmax=1.5, showscale=FALSE, height=500) %>%
  layout(title = "P_AAE10 Flat Heatmap", xaxis = list(title = "Clay %"), yaxis = list(title = "P_AAE10 (mg/kg)"), annotations = annotations_aae)

p4 <- plot_ly(x = 1:5, y = 1:26, z = z_aae10, type = "surface", colorscale = colorscale, cmin=0, cmax=1.5, showscale=FALSE, height=500) %>%
  layout(title = "P_AAE10 3D Staircase", scene = list(
    xaxis = list(title = "Clay %", tickvals=1:5, ticktext=x_clay),
    yaxis = list(title = "P_AAE10 (mg/kg)", tickvals=1:26, ticktext=y_aae10),
    zaxis = list(title = "Correction Factor C")
  ))

browsable(tagList(
  div(style="display:flex; justify-content:space-between;",
      div(style="width:48%;", p3),
      div(style="width:48%;", p4)
  )
))
```


---

## Interactive Bias Surface (Clay & pH) { .scrollable .smaller }

Here is a 3D visualization of the systematic bias ($\Delta v = v_{CO2} - v_{AAE10}$) modeled as a smooth surface (GAM) across the entire dataset. You can interact with the surface to see exactly how much the fertilizer recommendations diverge depending on **Clay** and **pH**!

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8
library(plotly)
library(mgcv)
library(tidyverse)

# Load data and define D_eval
final_artifacts <- readRDS("data/Final_Models_Data.rds")
D_Long_Agro <- final_artifacts$data$D_Long_Agro

get_clay_bin <- function(clay) {
  if (is.na(clay)) return(NA)
  if (clay < 10) return(1)
  if (clay < 20) return(2)
  if (clay < 30) return(3)
  if (clay < 40) return(4)
  return(5)
}
get_co2_bin <- function(p) {
  if (is.na(p)) return(NA)
  breaks <- seq(0.310, 4.650, by=0.310)
  idx <- findInterval(p, breaks) + 1
  return(min(idx, 16))
}
get_aae_bin <- function(p) {
  if (is.na(p)) return(NA)
  breaks <- seq(5.0, 125.0, by=5.0)
  idx <- findInterval(p, breaks) + 1
  return(min(idx, 16))
}

D_eval <- D_Long_Agro |> filter(!is.na(soil_0_20_P_CO2) & !is.na(soil_0_20_P_AAE10) & !is.na(rollMean_soil_0_20_clay) & !is.na(rollMean_soil_0_20_pH_H2O))
D_eval <- D_eval |> rowwise() |> mutate(
  clay_bin = get_clay_bin(rollMean_soil_0_20_clay),
  co2_bin = get_co2_bin(soil_0_20_P_CO2),
  aae_bin = get_aae_bin(soil_0_20_P_AAE10),
  v_co2 = z_co2[co2_bin, clay_bin],
  v_aae10 = z_aae10[aae_bin, clay_bin],
  bias = v_co2 - v_aae10
) |> ungroup()

# Fit a GAM to smooth the bias surface dynamically (now using pH instead of Corg)
fit <- gam(bias ~ s(rollMean_soil_0_20_clay, bs="cs") + s(rollMean_soil_0_20_pH_H2O, bs="cs"), data=D_eval)

clay_seq <- seq(min(D_eval$rollMean_soil_0_20_clay), max(D_eval$rollMean_soil_0_20_clay), length.out=50)
ph_seq <- seq(min(D_eval$rollMean_soil_0_20_pH_H2O), max(D_eval$rollMean_soil_0_20_pH_H2O), length.out=50)
grid <- expand.grid(rollMean_soil_0_20_clay = clay_seq, rollMean_soil_0_20_pH_H2O = ph_seq)
grid$pred_bias <- predict(fit, grid)

bias_matrix <- matrix(grid$pred_bias, nrow=50, ncol=50)

# Render Plotly Surface directly
plot_ly(x = clay_seq, y = ph_seq, z = bias_matrix, type = "surface", 
        colorscale = "RdBu", cmin = -1.5, cmax = 1.5) %>%
  layout(title = "Systematic Bias: v_CO2 - v_AAE10",
         scene = list(
           xaxis = list(title = "Clay %"),
           yaxis = list(title = "pH (H2O)"),
           zaxis = list(title = "Bias (CO2 - AAE10)")
         ))
```


---

## The Affine Morphism vs. Chemical Reality { .smaller }

**The Mathematical Implication of the GRUD Matrices**
Since we proved that both lookup tables are fundamentally affine maps targeting the identical output space ($v \in \mathbb{Q}$), equating them ($v_{CO2} = v_{AAE10}$) mathematically forces an affine morphism between the two extractions:

$$ \alpha_1 P_{CO2} + \beta_1 \text{clay} + \gamma_1 = \alpha_2 P_{AAE10} + \beta_2 \text{clay} + \gamma_2 $$

Solving for $P_{AAE10}$ yields a strict linear mapping:
$$ P_{AAE10} = c_1 \cdot P_{CO2} + c_2 \cdot \text{clay} + c_3 $$

**The Phenomenological Flaw**
This affine morphism mathematically dictates that the two extractions are linearly interchangeable. But chemically, they measure entirely different phenomena:
- **$P_{CO2}$** measures the immediately available concentration (**Intensity, $I$**).
- **$P_{AAE10}$** measures a much larger, strongly bound pool (**Quantity, $Q$**).

The true thermodynamic relationship between Quantity and Intensity is inherently non-linear (the Freundlich isotherm: $Q = K \cdot I^n$) and is heavily governed by **pH**. By forcing an affine bijection, the GRUD model entirely ignores the fact that at $pH > 7.3$, the EDTA chelator in AAE10 physically saturates from excess Calcium, causing the mathematical bijection to completely collapse.


---

## The Breakdown of Covariance: Calcareous Soils { .scrollable .smaller }

**The Observable Phenomenon**
We do not measure "True P". We measure the covariance between Intensity ($P_{CO2}$) and Quantity ($P_{AAE10}$). In non-calcareous soils (pH $\le$ 7.3), these two extractions co-vary predictably. However, when pH exceeds 7.3, this relationship violently breaks down: $P_{CO2}$ continues to extract a wide variance of phosphorus, while $P_{AAE10}$ collapses to near-zero.

**The Chemical Mechanism**
This collapse is not a deviation from some unmeasurable truth; it is a physical failure of the AAE10 method itself. AAE10 uses an ammonium acetate-EDTA buffer calibrated to pH 4.65. In highly calcareous soils, the massive presence of free $CaCO_3$ instantly neutralizes the acid buffer, and the flood of $Ca^{2+}$ ions physically saturates the EDTA chelator. The affine mathematical model forced by the GRUD completely ignores this chemical boundary.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| fig-width: 14
#| fig-height: 6

D_plot <- D_Long_Agro |> 
  filter(!is.na(soil_0_20_P_CO2) & !is.na(soil_0_20_P_AAE10) & !is.na(rollMean_soil_0_20_pH_H2O)) |>
  mutate(is_high_ph = ifelse(rollMean_soil_0_20_pH_H2O > 7.3, "pH > 7.3 (Calcareous)", "pH <= 7.3"))

ggplot(D_plot, aes(x = soil_0_20_P_CO2, y = soil_0_20_P_AAE10, color = rollMean_soil_0_20_pH_H2O)) +
  geom_point(alpha = 0.6, size=2) +
  scale_color_viridis_c(option = "magma", name = "Soil pH") +
  geom_smooth(method="gam", color="blue", fill="blue", alpha=0.2) +
  facet_wrap(~is_high_ph, scales="free_y") +
  labs(
    title = "The Breakdown of Covariance in Calcareous Soils",
    subtitle = "Notice how the relationship collapses violently above pH 7.3",
    x = "Phosphorus Intensity (CO2 Extraction, mg/L)",
    y = "Phosphorus Quantity (AAE10 Extraction, mg/kg)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "bottom"
  )
```

---

## The Illusion of Precision: Analytical Noise { .smaller }

The GRUD projects a false sense of administrative precision by forcing rigid class boundaries that the laboratory hardware itself cannot reliably support.

**$P_{AAE10}$ Method Accuracy vs. Policy Boundaries**
- **GRUD Bin Width:** 5.0 mg P/kg
- **Lab Standard Deviation:** 1.0 mg P/kg (95% CI = $\pm$ 2.0 mg P/kg)
- **The Reality:** The 95% confidence interval spans **4.0 mg P/kg**, consuming **80% of the entire classification bin**. A sample with a true value of 9.0 mg P/kg (Class C) could easily measure as 7.0 (Class C) or 11.0 (Class B) strictly due to photometer noise.

**$P_{CO2}$ Method Accuracy vs. Policy Boundaries**
- **GRUD Bin Width:** 0.31 mg P/kg (exactly 2.0 P-Testzahlen)
- **Lab Standard Deviation:** up to ~0.36 Testzahlen (95% CI = $\pm$ 0.72 Testzahlen)
- **The Reality:** The 95% confidence interval spans **~1.4 Testzahlen**, consuming **70% of the bin width**. 

By imposing strict step-functions, the GRUD forces farmers over arbitrary fertilizer cliffs based on variance that is entirely within the margin of error of the photometer. A continuous mechanistic curve naturally absorbs this noise.

